---
title: "Lundie-Brown et al 2026"
output: html_notebook
---

This code is for producing the publication figures in Lundie-Brown et al., 2026

Preprocessing objects are found in other R markdown files and details of required output objects are given here: 


RNAseq experiment 1: 
 
  RNAseq_ASCL1_dds.Rds - DESeq2 object 
  RNAseq_ASCL1_DEGs.csv - sheet of degs and which condtion / timepoint they are upregulated in 
  RNAseq_ASCL1_zmtrx.txt - z-scored expression matrix 

ChIPseq experiment 1: 
ATACseq experiment 1: 

  ATACseq_analysed_RLE.Rds - analysed diffbind object for the ATAC-seq data 
  
RNAseq experiment 2: 
RNAseq experiment 3: 


# FIG 1 

Fig 1F - PCA plot

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(PCAtools)

# load in the dds object

dds <- readRDS(here("../Data/RNAseq_ASCL1_dds.Rds"))

# calculating the pca plot and variance stabilised data

rna_vsd <- assay(vst(dds, blind=FALSE))

rna_pca <- pca(rna_vsd, metadata = colData(dds), removeVar = 0.95) # takes just top 5% variable genes 

# plot pca 1 versus 2 (Figu 2B) based on top 5% of variable genes

rna_pcaRot <- rna_pca$rotated[,c(1:3)]
rna_pcaRot <- cbind(rna_pcaRot, Group = rna_pca$metadata$Group)
rna_pcaPcntVar <- round(rna_pca$variance)

colours <- list("C11_mESC_t24" = "lightgrey", "C11_EpiLC_t24" = "lightgrey", "C11_NE_t24" = "lightgrey",
                "gG5_mESC_t0" = "#AAE4E9", "gG5_mESC_t6" = "#79C9D1", "gG5_mESC_t24" = "#4FAAB2",
                "gG5_EpiLC_t0" = "#FFE8B9", "gG5_EpiLC_t6" = "#FFDC92", "gG5_EpiLC_t24" = "#FFD170",
                "gG5_NE_t0" = "#FFBCB9", "gG5_NE_t6" = "#FF9792", "gG5_NE_t24" = "#FF7770")


rna_pcaFig12 <- ggplot(rna_pcaRot, aes(PC1, PC2, colour = Group)) + 
  geom_point(size = 3) +
  xlab(paste0("PC1: ",rna_pcaPcntVar[1],"% variance")) +
  ylab(paste0("PC2: ",rna_pcaPcntVar[2],"% variance")) +
  scale_colour_manual(values = unlist(colours), guide = "none") + 
  theme_bw()

rna_pcaFig12

png(here("../Figures/Fig1F.png"), width = 1000, height = 1000, res = 300)
print(rna_pcaFig12)
dev.off()


# rna_pcaFig23 <- ggplot(rna_pcaRot, aes(PC2, PC3, colour = Group)) + 
#   geom_point(size = 3) +
#   xlab(paste0("PC2: ",rna_pcaPcntVar[2],"% variance")) +
#   ylab(paste0("PC3: ",rna_pcaPcntVar[3],"% variance")) +
#   scale_colour_manual(values = unlist(colours), guide = "none") +
#   theme_bw()

# png(here("../Figures/FigS2A.png"), width = 1000, height = 1000, res = 300)
# print(rna_pcaFig23)
# dev.off()

```

Fig 1G - how many DEGs after ASCL1 in each cell type at 6 and 24 hours

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(ggplot2)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))


plotdat <- data.frame(
  cond = factor(c("mESC_t6", "mESC_t24", "EpiLC_t6", "EpiLC_t24", "NE_t6", "NE_t24"), levels = c("mESC_t6", "mESC_t24", "EpiLC_t6", "EpiLC_t24", "NE_t6", "NE_t24")),
  up = c(sum(degs$mESC_t6_gg5_change == "up"),
         sum(degs$mESC_t24_gg5_change == "up"),
         sum(degs$EpiLC_t6_gg5_change == "up"),
         sum(degs$EpiLC_t24_gg5_change == "up"),
         sum(degs$NE_t6_gg5_change == "up"),
         sum(degs$NE_t24_gg5_change == "up")),
  down = c(sum(degs$mESC_t6_gg5_change == "down"),
         sum(degs$mESC_t24_gg5_change == "down"),
         sum(degs$EpiLC_t6_gg5_change == "down"),
         sum(degs$EpiLC_t24_gg5_change == "down"),
         sum(degs$NE_t6_gg5_change == "down"),
         sum(degs$NE_t24_gg5_change == "down")))


plotdat_long <- plotdat %>%
  tidyr::pivot_longer(cols = c(up, down), names_to = "direction", values_to = "count") %>%
  dplyr::mutate(count = ifelse(direction == "down", -count, count))


p = ggplot(plotdat_long, aes(x = cond, y = count, fill = cond, group = direction)) +
  geom_col(show.legend = FALSE, position = position_dodge(width = 0.9), width = 0.7) +
  geom_hline(yintercept = 0, colour = "black") +
  scale_fill_manual(values = c("mESC_t6" = "#76C9D1", 
                               "mESC_t24" = "#4FAAB2", 
                               "EpiLC_t6" = "#FFDC92", 
                               "EpiLC_t24" = "#FFD170", 
                               "NE_t6" = "#FF9792", 
                               "NE_t24" = "#FF7770")) +
  labs(x = NULL, y = "Number of DEGs") +
  theme_bw() +
  theme(axis.text.x = element_blank())

p

ggsave(plot = p, filename = here("../Figures/Fig1G.png"), height = 4, width = 4, dpi = 300)

```

Fig 1H - venn diagram of DEGs up at 24 hours 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(eulerr)

# venn diagram 24 hour 

euler_24h <- eulerr::euler(list(
  "mESC 24h Dex" = degs %>% dplyr::filter(mESC_t24_gg5_change == "up") %>% pull(symbol),
  "EpiLC 24h Dex" = degs %>% dplyr::filter(EpiLC_t24_gg5_change == "up") %>% pull(symbol),
  "NE 24h Dex" = degs %>% dplyr::filter(NE_t24_gg5_change == "up") %>% pull(symbol)
))


colours_t24 <- c("#4FAAB2", "#FFD170", "#FF7770")


t24 <- plot(euler_24h,
            fill = colours_t24,
            edges = list(col = "black"),
            quantities = TRUE)

t24

ggsave(here("../Figures/Fig1H.png"), t24, width = 4, height = 4, dpi = 300)

```

Fig 1I - heatmap of all upregulated DEGs across conditions 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DESeq2)
library(ComplexHeatmap)
library(MetBrewer)

# colours for plotting 

colours <- c("C11_mESC_t24" = "lightgrey", "C11_EpiLC_t24" = "lightgrey", "C11_NE_t24" = "lightgrey",
             "gG5_mESC_t0" = "#AAE4E9","gG5_mESC_t6" = "#79C9D1","gG5_mESC_t24" = "#4FAAB2",
             "gG5_EpiLC_t0" = "#FFE8B9","gG5_EpiLC_t6" = "#FFDC92","gG5_EpiLC_t24" = "#FFD170",
             "gG5_NE_t0" = "#FFBCB9","gG5_NE_t6" = "#FF9792","gG5_NE_t24" = "#FF7770")


# load in the z-scored matrix 

vst.zmtrx <- read.table(file = here("../Data/RNAseq_ASCL1_zmtrx.txt"), row.names = 1) %>% as.matrix()

# determine a list of genes to plot from the total deg list and filter the heatmap matrix 

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

degs_t24 <- degs %>% dplyr::filter(mESC_t24_gg5_change == "up" | EpiLC_t24_gg5_change == "up" | NE_t24_gg5_change == "up") %>% pull(symbol)

vst.zmtrx.24degs <- vst.zmtrx[rownames(vst.zmtrx) %in% degs_t24, ]


# prepare annotation using colnames and dds object 

dds <- readRDS(here("../Data/RNAseq_ASCL1_dds.Rds"))

col_annotation <- colData(dds)[colnames(vst.zmtrx), "Group"]

heatmap_annotation_head <- HeatmapAnnotation(Condition = col_annotation,
                                             col = list(Condition = colours),
                                             show_legend = FALSE,
                                             show_annotation_name = FALSE
                                             )

# prepare a row split factor 

split_order <- c("all_three", "mesc_only", "mesc_epilc", "epilc_only", "epilc_ne", "ne_only", "mesc_ne")

row_split <- factor(
  degs$t24up_group[match(rownames(vst.zmtrx.24degs), degs$symbol)],
  levels = split_order
)

# Count how many genes are in each category
split_counts <- table(row_split)

# Create new labels like "only_mESC (37 genes)"
split_labels <- paste0(names(split_counts), "\n (", split_counts, " genes)")

# Assign the labels as levels
levels(row_split) <- split_labels


# make gene name annotations 

slct_genes <- c("Tnni2", "Vgll1", "Esx1", "Dppa1", "Cebpa", "Cebpb", "Sync", "Eno3", "Dll3", "Id1", "Id3", "Hes5",
                "Plxna2", "Wnt5a", "Tubb3", "Map2", "Isl1", "Phox2a", "Myt1", "Elavl4", "Ank2", "Gsx2", "Sox10", "Sox11", "Sox3", "Onecut2", "Onecut1", "Dcx")


mark_index <- which(rownames(vst.zmtrx.24degs) %in% slct_genes)
mark_labels <- rownames(vst.zmtrx.24degs)[mark_index]

right_annot <- rowAnnotation(
  gene = anno_mark(
    at = mark_index,
    labels = mark_labels,
    labels_gp = gpar(fontsize = 20),
    link_width = unit(8, "mm"),       # distance between heatmap & labels
    padding = unit(1, "mm")
  )
)

# Generate the heatmap
heatmap_obj <- Heatmap(
  name = "z-scored expression",
  vst.zmtrx.24degs,
  col = met.brewer("OKeeffe1", direction = -1),
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  top_annotation = heatmap_annotation_head,
  row_split = row_split,
  row_title_gp = gpar(fontsize = 5, fontface = "bold"),
  cluster_row_slices = FALSE,
  heatmap_legend_param = list(legend_direction = "horizontal")
)

# add text labels 
heatmap_wlabels <- heatmap_obj + right_annot

draw(heatmap_wlabels, heatmap_legend_side = "bottom")

# export the plot 
png(file = here("../Figures/Fig1I.png"), width = 6, height = 10, units = "in", res = 300)
draw(heatmap_wlabels, heatmap_legend_side = "bottom")
dev.off()

```

Fig 1J–L - GO analysis and plots of upregulated DEGs at 24 hours 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(ggplot2)
library(clusterProfiler)
library(org.Mm.eg.db)
library(cowplot)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

# plot top 5 enriched (& simplified) GO terms for upregulated genes at 24 hours in each cell type - not clustered, so some genes will be shared between conditions 

mesc_up <- degs %>%
  dplyr::filter(mESC_t24_gg5_change == "up") %>%
  pull(symbol)

go_mesc <- enrichGO(
  gene          = mesc_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_mesc <- simplify(go_mesc)

p_mesc <- barplot(smp_mesc, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_fill_gradient(high = "#AAE4E9", low = "#79C9D1")

p_mesc

ggsave(plot = p_mesc, filename = here("../Figures/Fig1J-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")
  


epilc_up <- degs %>%
  dplyr::filter(EpiLC_t24_gg5_change == "up") %>%
  pull(symbol)

go_epilc <- enrichGO(
  gene          = epilc_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_epilc <- simplify(go_epilc)

p_epilc <- barplot(smp_epilc, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 18)) +
  scale_fill_gradient(high = "#FFE8B9", low = "#FFDC92")

p_epilc

ggsave(plot = p_epilc, filename = here("../Figures/Fig1K-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")



ne_up <- degs %>%
  dplyr::filter(NE_t24_gg5_change == "up") %>%
  pull(symbol)

go_ne <- enrichGO(
  gene          = ne_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_ne <- simplify(go_ne)

p_ne <- barplot(smp_ne, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_fill_gradient(high = "#FFBCB9", low = "#FF9792")

p_ne

ggsave(plot = p_ne, filename = here("../Figures/Fig1L-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")

```


# FIG S1

Fig S1I–K - volcano plots of DEGs after 24 hours ASCL1 in each condition 

```{r, messages=FALSE}

# note the log fold change is shrunk using ashr, represents the comparison against t0 gg5, but changes are filtered to be also significant in the vs c11 comparison 

library(here)
library(tidyverse)
library(ggplot2)
library(ggrepel)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

#----------------------------------#
# define the volcano plot function #
#----------------------------------#

plotVolcano <- function(degs, contrast_prefix,
                        x_lim = c(-10, 10),
                        y_lim = c(0, 80),
                        upcolour = "#76C9D1",
                        nscolour = "grey80",
                        downcolour = "grey30",
                        top_n = 10) {
  
  # build column names based on prefix
  lfc_col <- paste0(contrast_prefix, "_ashrLFC")
  padj_col <- paste0(contrast_prefix, "_padj")
  change_col <- paste0(contrast_prefix, "_change")
  
  # prepare plot data
  plotdat <- degs %>%
    dplyr::mutate(
      lfc = !!rlang::sym(lfc_col),
      log_padj = -log10(!!rlang::sym(padj_col)),
      change = factor(!!rlang::sym(change_col), levels = c("up", "ns", "down")),
      oob_x = lfc < x_lim[1] | lfc > x_lim[2],
      oob_y = log_padj < y_lim[1] | log_padj > y_lim[2]
    ) %>%
    dplyr::select(symbol, lfc, log_padj, change, oob_x, oob_y)
  
  # top N inside-bound genes
  top_genes <- plotdat %>%
    dplyr::filter(!(oob_x | oob_y)) %>%
    dplyr::arrange(desc(abs(log_padj))) %>%
    dplyr::slice_head(n = top_n)
  
  # combine labels for top genes + out-of-bounds
  labels <- bind_rows(
    top_genes %>%
      dplyr::mutate(lfc_capped = lfc, log_padj_capped = log_padj),
    plotdat %>%
      dplyr::filter(oob_x | oob_y) %>%
      dplyr::mutate(lfc_capped = pmin(pmax(lfc, x_lim[1]), x_lim[2]),
                    log_padj_capped = pmin(pmax(log_padj, y_lim[1]), y_lim[2]))
  )
  
  # make the volcano plot
  ggplot(plotdat, aes(x = lfc, y = log_padj, colour = change)) +
    geom_point(data = dplyr::filter(plotdat, !(oob_x | oob_y)),
               size = 2, alpha = 0.6, show.legend = FALSE) +
    geom_point(data = dplyr::filter(plotdat, oob_x | oob_y),
               aes(x = pmin(pmax(lfc, x_lim[1]), x_lim[2]),
                   y = pmin(pmax(log_padj, y_lim[1]), y_lim[2])),
               size = 2, shape = 6, show.legend = FALSE) +
    geom_text_repel(data = labels,
                    aes(x = lfc_capped, y = log_padj_capped, label = symbol),
                    box.padding = 0.8, max.overlaps = Inf, size = 4, colour = "black") +
    geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "black") +
    coord_cartesian(xlim = x_lim, ylim = y_lim, expand = FALSE) +
    labs(y = expression("−log"["10"] ~ "adjusted P value"),
         x = expression("log"["2"] ~ "fold change (ashr)")) +
  scale_colour_manual(values = c(up = upcolour, ns = nscolour, down = downcolour)) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
}

plotVolcano(degs, "mESC_t24_gg5", top_n = 2)
plotVolcano(degs, "EpiLC_t24_gg5", upcolour = "#FDD163")
plotVolcano(degs, "NE_t24_gg5", upcolour = "#FB7D80")

#----------------------------------#
# using the volcano plot function #
#----------------------------------#

ggsave(plot = plotVolcano(degs, "mESC_t24_gg5", top_n = 0), file = here("../Figures/Fig1J.png"), height = 4, width = 3.5, dpi = 300)
ggsave(plot = plotVolcano(degs, "EpiLC_t24_gg5", upcolour = "#FDD163"), file = here("../Figures/Fig1K.png"), height = 4, width = 3.5,, dpi = 300)
ggsave(plot = plotVolcano(degs, "NE_t24_gg5", upcolour = "#FB7D80"), file = here("../Figures/Fig1L.png"), height = 4, width = 3.5,, dpi = 300)

```

Fig S1L - venn diagram of DEGs up at 6 hours 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(ggplot2)
library(eulerr)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

# venn diagram 6 hour

euler_6h <- eulerr::euler(list(
  "mESC 6h Dex" = degs %>% dplyr::filter(mESC_t6_gg5_change == "up") %>% pull(symbol),
  "EpiLC 6h Dex" = degs %>% dplyr::filter(EpiLC_t6_gg5_change == "up") %>% pull(symbol),
  "NE 6h Dex" = degs %>% dplyr::filter(NE_t6_gg5_change == "up") %>% pull(symbol)
))

colours_t6 <- c("#79C9D1", "#FFDC92", "#FF9792")

t6 <- plot(euler_6h,
           fill = colours_t6,
           edges = list(col = "black"),
           #labels = list(font = 1, cex = 1.2, col = "black", labels = custom_labels),
           quantities = TRUE
)

t6 

ggsave(here("../Figures/FigS1I.png"), t6, width = 4, height = 4, dpi = 300)

```

Fig S1M-O - volcano plots of DEGs after 6 hours 

```{r, messages=FALSE}

# note the log fold change is shrunk using ashr, represents the comparison against t0 gg5, but changes are filtered to be also significant in the vs c11 comparison 

library(here)
library(tidyverse)
library(ggplot2)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

#----------------------------------#
# define the volcano plot function #
#----------------------------------#

library(dplyr)
library(ggplot2)
library(ggrepel)

plotVolcano <- function(degs, contrast_prefix,
                        x_lim = c(-10, 10),
                        y_lim = c(0, 80),
                        upcolour = "#76C9D1",
                        nscolour = "grey80",
                        downcolour = "grey30",
                        top_n = 10) {
  
  # build column names based on prefix
  lfc_col <- paste0(contrast_prefix, "_ashrLFC")
  padj_col <- paste0(contrast_prefix, "_padj")
  change_col <- paste0(contrast_prefix, "_change")
  
  # prepare plot data
  plotdat <- degs %>%
    dplyr::mutate(
      lfc = !!rlang::sym(lfc_col),
      log_padj = -log10(!!rlang::sym(padj_col)),
      change = factor(!!rlang::sym(change_col), levels = c("up", "ns", "down")),
      oob_x = lfc < x_lim[1] | lfc > x_lim[2],
      oob_y = log_padj < y_lim[1] | log_padj > y_lim[2]
    ) %>%
    dplyr::select(symbol, lfc, log_padj, change, oob_x, oob_y)
  
  # top N inside-bound genes
  top_genes <- plotdat %>%
    dplyr::filter(!(oob_x | oob_y)) %>%
    dplyr::arrange(desc(abs(log_padj))) %>%
    dplyr::slice_head(n = top_n)
  
  # combine labels for top genes + out-of-bounds
  labels <- bind_rows(
    top_genes %>%
      dplyr::mutate(lfc_capped = lfc, log_padj_capped = log_padj),
    plotdat %>%
      dplyr::filter(oob_x | oob_y) %>%
      dplyr::mutate(lfc_capped = pmin(pmax(lfc, x_lim[1]), x_lim[2]),
                    log_padj_capped = pmin(pmax(log_padj, y_lim[1]), y_lim[2]))
  )
  
  # make the volcano plot
  ggplot(plotdat, aes(x = lfc, y = log_padj, colour = change)) +
    geom_point(data = dplyr::filter(plotdat, !(oob_x | oob_y)),
               size = 2, alpha = 0.6, show.legend = FALSE) +
    geom_point(data = dplyr::filter(plotdat, oob_x | oob_y),
               aes(x = pmin(pmax(lfc, x_lim[1]), x_lim[2]),
                   y = pmin(pmax(log_padj, y_lim[1]), y_lim[2])),
               size = 2, shape = 6, show.legend = FALSE) +
    geom_text_repel(data = labels,
                    aes(x = lfc_capped, y = log_padj_capped, label = symbol),
                    box.padding = 0.8, max.overlaps = Inf, size = 4, colour = "black") +
    geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "black") +
    coord_cartesian(xlim = x_lim, ylim = y_lim, expand = FALSE) +
    labs(y = expression("−log"["10"] ~ "adjusted P value"),
         x = expression("log"["2"] ~ "fold change (ashr)")) +
  scale_colour_manual(values = c(up = upcolour, ns = nscolour, down = downcolour)) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
}

p1 <- plotVolcano(degs, "mESC_t6_gg5", top_n = 10)
p1

p2 <- plotVolcano(degs, "EpiLC_t6_gg5", upcolour = "#FDD163")
p2

p3 <- plotVolcano(degs, "NE_t6_gg5", upcolour = "#FB7D80")
p3

ggsave(plot = p1, file = here("../Figures/FigS1J-1.png"), height = 4, width = 3.5, dpi = 300)
ggsave(plot = p2, file = here("../Figures/FigS1K-1.png"), height = 4, width = 3.5,, dpi = 300)
ggsave(plot = p3, file = here("../Figures/FigS1L-1.png"), height = 4, width = 3.5, dpi = 300)

```

Fig S1M-O - GO analysis of DEGs after 6 hours to match volcano plots 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(ggplot2)
library(clusterProfiler)
library(org.Mm.eg.db)
library(cowplot)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

# plot top 5 enriched (& simplified) GO terms for upregulated genes at 24 hours in each cell type - not clustered, so some genes will be shared between conditions 

mesc_up <- degs %>%
  dplyr::filter(mESC_t6_gg5_change == "up") %>%
  pull(symbol)

go_mesc <- enrichGO(
  gene          = mesc_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_mesc <- simplify(go_mesc)

p_mesc <- barplot(smp_mesc, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_fill_gradient(high = "#AAE4E9", low = "#79C9D1")

p_mesc

ggsave(plot = p_mesc, filename = here("../Figures/FigS1J-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")



epilc_up <- degs %>%
  dplyr::filter(EpiLC_t6_gg5_change == "up") %>%
  pull(symbol)

go_epilc <- enrichGO(
  gene          = epilc_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_epilc <- simplify(go_epilc)

p_epilc <- barplot(smp_epilc, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_fill_gradient(high = "#FFE8B9", low = "#FFDC92")

p_epilc

ggsave(plot = p_epilc, filename = here("../Figures/FigS1K-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")



ne_up <- degs %>%
  dplyr::filter(NE_t6_gg5_change == "up") %>%
  pull(symbol)

go_ne <- enrichGO(
  gene          = ne_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",    
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2)

smp_ne <- simplify(go_ne)

p_ne <- barplot(smp_ne, showCategory = 5) + theme_cowplot() + scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_fill_gradient(high = "#FFBCB9", low = "#FF9792")

p_ne

ggsave(plot = p_ne, filename = here("../Figures/FigS1L-2.png"), width = 3.5, height = 4, dpi = 300, bg = "white")

```


# FIG 2 

Fig 2A - deeptools heatmap for ASCL1 ChIP-seq split by cell type 

```{bash}

computeMatrix scale-regions -S \
bigwigs/gG5_mESC_HA_RLE_merged.bigwig \
bigwigs/gG5_EpiLC_HA_RLE_merged.bigwig \
bigwigs/gG5_NE_HA_RLE_merged.bigwig \
-R beds/ChIPseq_ASCL1_mesconly10k.bed \
beds/ChIPseq_ASCL1_mescepilc10k.bed \
beds/ChIPseq_ASCL1_epilconly10k.bed \
beds/ChIPseq_ASCL1_epilcne10k.bed \
beds/ChIPseq_ASCL1_neonly10k.bed \
beds/ChIPseq_ASCL1_mescne10k.bed \
beds/ChIPseq_ASCL1_universal10k.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_ascl1chip.mat.gz
 
plotHeatmap -m matrix/matrix_ascl1chip.mat.gz -out plots/ascl1_chip.png --whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ASCL1" "ASCL1" --startLabel "" --endLabel "" --colorMap Greys Greys Greys --zMin 0 0 0 --zMax 100 100 100 --sortRegions descend

```

Fig 2B-C - specific gene counts for Myt1 and Hdac11

```{r}
library(here)
library(AnnotationHub)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(cowplot)

# load in the rna-seq rds object
# these counts are corrected for library size and Tx length
rna <- readRDS(file = here("../Data/RNAseq_ASCL1_dds.Rds"))
rna <- estimateSizeFactors(rna)
rnaseq_counts <- counts(object = rna, normalized = TRUE)
# 
# # write to a csv 
# write.csv(rnaseq_counts, file = here("../results/countsmatrix_normalised.csv"))

# need to annotate the row names to gene symbols
ah <- AnnotationHub()
# ah object ID for the mm39 108 ensembl build is AH109367
EnsDb.Mmusculus.v108 <- ah[["AH109367"]]
genes <- genes(EnsDb.Mmusculus.v108, return.type = "DataFrame")
gene2symbol <- genes[c(1,12)]
gene2symbol_matched <- gene2symbol$symbol[match(rownames(rnaseq_counts), gene2symbol$gene_id)]
# annotate the rna-seq table
rownames(rnaseq_counts) <- gene2symbol_matched

colours <- c("C11_mESC_t24" = "lightgrey", "C11_EpiLC_t24" = "lightgrey", "C11_NE_t24" = "lightgrey",
             "gG5_mESC_t0" = "#AAE4E9","gG5_mESC_t6" = "#79C9D1","gG5_mESC_t24" = "#4FAAB2",
             "gG5_EpiLC_t0" = "#FFE8B9","gG5_EpiLC_t6" = "#FFDC92","gG5_EpiLC_t24" = "#FFD170",
             "gG5_NE_t0" = "#FFBCB9","gG5_NE_t6" = "#FF9792","gG5_NE_t24" = "#FF7770")

# before generating the heatmap, want to change the order so that the C11 columns are next to the t0 columns 
desired_order <- c(
  # mESC
  "exp1_C11_mESC_t24_rep1", "exp1_C11_mESC_t24_rep2", "exp1_C11_mESC_t24_rep3",
  "exp2_gG5_mESC_t0_rep1", "exp2_gG5_mESC_t0_rep2", "exp2_gG5_mESC_t0_rep3", "exp2_gG5_mESC_t0_rep4",
  "exp2_gG5_mESC_t6_rep1", "exp2_gG5_mESC_t6_rep2", "exp2_gG5_mESC_t6_rep3", "exp2_gG5_mESC_t6_rep4",
  "exp1_gG5_mESC_t24_rep1", "exp1_gG5_mESC_t24_rep2", "exp1_gG5_mESC_t24_rep3", 
  "exp2_gG5_mESC_t24_rep1", "exp2_gG5_mESC_t24_rep2", "exp2_gG5_mESC_t24_rep3", "exp2_gG5_mESC_t24_rep4",
  
  # EpiLC
  "exp1_C11_EpiLC_t24_rep1", "exp1_C11_EpiLC_t24_rep2", "exp1_C11_EpiLC_t24_rep3",
  "exp2_gG5_EpiLC_t0_rep1", "exp2_gG5_EpiLC_t0_rep2", "exp2_gG5_EpiLC_t0_rep3", "exp2_gG5_EpiLC_t0_rep4",
  "exp2_gG5_EpiLC_t6_rep1", "exp2_gG5_EpiLC_t6_rep2", "exp2_gG5_EpiLC_t6_rep3", "exp2_gG5_EpiLC_t6_rep4",
  "exp1_gG5_EpiLC_t24_rep1", "exp1_gG5_EpiLC_t24_rep2", "exp1_gG5_EpiLC_t24_rep3", 
  "exp2_gG5_EpiLC_t24_rep1", "exp2_gG5_EpiLC_t24_rep2", "exp2_gG5_EpiLC_t24_rep3",
  
  # NE
  "exp1_C11_NE_t24_rep1", "exp1_C11_NE_t24_rep2", "exp1_C11_NE_t24_rep3", 
  "exp2_gG5_NE_t0_rep1", "exp2_gG5_NE_t0_rep2", "exp2_gG5_NE_t0_rep3", "exp2_gG5_NE_t0_rep4",
  "exp2_gG5_NE_t6_rep1", "exp2_gG5_NE_t6_rep2", "exp2_gG5_NE_t6_rep3", "exp2_gG5_NE_t6_rep4",
  "exp1_gG5_NE_t24_rep1", "exp1_gG5_NE_t24_rep2", "exp1_gG5_NE_t24_rep3", 
  "exp2_gG5_NE_t24_rep1", "exp2_gG5_NE_t24_rep2", "exp2_gG5_NE_t24_rep3", "exp2_gG5_NE_t24_rep4"
)


# Reorder the heatmap data and annotation
rnaseq_counts <- rnaseq_counts[, desired_order]

genes.ofi <- c("Hdac11", "Myt1")

# Loop through each gene and create + save the plot
for (gene.ofi in genes.ofi) {
  
  # Get counts for current gene
  log.counts.ofi <- log2(rnaseq_counts[gene.ofi, ] + 1)
  counts.ofi <- rnaseq_counts[gene.ofi, ]
  
  # Create plot
  gg.logtpmBox <- counts.ofi %>%
    tibble::as_tibble() %>%
    mutate(sample = names(counts.ofi)) %>%
    separate(sample, c('exp', 'line', 'cell', 'time', 'rep'), sep = '_') %>%
    mutate(cond_time = paste(line, cell, time, sep = "_")) %>%
    mutate(cond_time = factor(cond_time, levels = unique(cond_time))) %>%
    ggplot(aes(x = cond_time, y = value, fill = cond_time)) +
    geom_boxplot(width = 0.7, outlier.shape = NA, color = "black", size = 0.3) +
    geom_point(
      aes(fill = cond_time),
      position = position_jitter(width = 0.1, height = 0),
      size = 2, shape = 21, color = "black"
    ) +
    xlab('') +
    #ylab(bquote(log[2]~"(scaled counts + 1)")) +
    ylab("scaled counts") +
    scale_fill_manual(values = colours) +
    scale_color_manual(values = colours) +
    scale_y_continuous(limits = c(NA, NA)) +
    theme_cowplot() +
    labs(title = gene.ofi) +
    theme(
      legend.position = 'none',
      axis.line.x = element_line(linewidth = 0.4),
      axis.line.y = element_line(linewidth = 0.4),
      axis.line = element_line(linetype = 'solid'),
      axis.text.x = element_blank(),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 16, face = "italic", hjust = 0.5)
    )
  
  # Save the plot
  ggsave(
    plot = gg.logtpmBox,
    filename = here("../Figures/", paste0("RNAseq_", gene.ofi, ".png")),
    device = "png", units = "cm", width = 5.5, height = 8, bg = "white"
  )
}
```

Fig 2D - categorising peaks based on cell type binding & proximal DEGs

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(GenomicRanges)

# which cell types are the peaks bound in 

degs <- read.table(file = here("../Data/RNAseq_ASCL1_DEGs.csv"), sep = ",", header = TRUE)

degs.neonly <- degs %>% dplyr::filter(t24up_group == "ne_only") # 235 genes in this list 

peaktogenes <- read.table(file = here("../Data/ChIPseq_allpeaktogenes.txt"), sep = "\t", header = TRUE,row.names = 1)

peaktogenes.neonly <- peaktogenes %>% 
  dplyr::filter(gene_symbol %in% dplyr::pull(degs.neonly, var = symbol)) %>%
  dplyr::filter(peakset %in% c("NE", "EpiLC_NE", "mESC_NE", "Universal"))

boundgenes <- peaktogenes.neonly %>% pull(gene_symbol) %>% unique() # 208 genes 

write.table(boundgenes, file = here("../Data/ChIPseq_NEboundgenes_208.txt"), quote = FALSE)

# find just the peaks 

peaks <- peaktogenes.neonly %>%
  dplyr::select(-gene_id, -gene_symbol) %>% 
  unique() %>% 
  GRanges() %>% 
  as.data.frame()

# this gives the 653 sites that we are interested in 

# write this to a BED file for further plotting 

write.table(peaks, file = here("../Data/ChIPseq_653NEsites.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# 198 are NE only 
write.table(peaks %>% dplyr::filter(peakset == "NE"), 
            file = here("../Data/ChIPseq_653NEsites_splitNEonly.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# 455 are not NE only 
write.table(peaks %>% dplyr::filter(peakset != "NE"), 
            file = here("../Data/ChIPseq_653NEsites_splitnotNEonly.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# 417 are not in mESC
write.table(peaks %>% dplyr::filter(!peakset %in% c("Universal", "mESC_NE")), 
            file = here("../Data/ChIPseq_653NEsites_splitnotmesc.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# 236 are in mESC
write.table(peaks %>% dplyr::filter(peakset %in% c("Universal", "mESC_NE")), 
            file = here("../Data/ChIPseq_653NEsites_splitmesc.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

```

Fig 2E - boxplots to show ChIP-seq signal strength at sites linked to NE-specific DEGs 

```{r, messages=FALSE}

library(rtracklayer)
library(GenomicRanges)
library(here)
library(tidyverse)
library(rstatix)
library(ggpubr)


ascl1peaks <- read.table(
  file = here("../Data/ChIPseq_NEspecificDEG_peaks_full.csv"),
  sep = ",",
  header = TRUE
)

neonly <- ascl1peaks %>%
  dplyr::filter(in_mESC_10k == FALSE & in_EpiLC_10k == FALSE) %>%
  dplyr::select(2:4) %>%
  GRanges() %>%
  unique()

notneonly <- ascl1peaks %>%
  dplyr::filter(in_mESC_10k == TRUE | in_EpiLC_10k == TRUE) %>%
  dplyr::select(2:4) %>%
  GRanges() %>%
  unique()

# nemesc <- ascl1peaks %>%
#   dplyr::filter(in_mESC_10k == TRUE) %>%
#   dplyr::select(2:4) %>%
#   GRanges() %>%
#   unique()
# 
# neepilc <- ascl1peaks %>%
#   dplyr::filter(in_EpiLC_10k == TRUE) %>%
#   dplyr::select(2:4) %>%
#   GRanges() %>%
#   unique()


plot_ascl1_signal <- function(
  regions_gr,
  label,
  outdir = here("../Figures")
) {

  bw_dir <- here("../Data/chipseq_ascl1-ha/bigwig/")

  bw_files <- normalizePath(
    list.files(
      bw_dir,
      pattern = "_merged\\.(bw|bigwig)$",
      full.names = TRUE
    ),
    winslash = "/",
    mustWork = TRUE
  )

  stopifnot(length(bw_files) == 3)

  extract_bw_sum <- function(bw_file, regions) {
    bw <- BigWigFile(bw_file)
    signal <- import(bw, which = regions)

    hits <- findOverlaps(signal, regions)
    if (length(hits) == 0) return(rep(0, length(regions)))

    ov <- pintersect(signal[queryHits(hits)], regions[subjectHits(hits)])
    weighted_signal <- signal$score[queryHits(hits)] * width(ov)
    region_sum <- tapply(weighted_signal, subjectHits(hits), sum)

    out <- numeric(length(regions))
    out[as.integer(names(region_sum))] <- region_sum
    out
  }

  signal_matrix <- sapply(bw_files, extract_bw_sum, regions = regions_gr)
  colnames(signal_matrix) <- sub("\\.bigwig$|\\.bw$", "", basename(bw_files))

  df_long <- as.data.frame(signal_matrix) %>%
    mutate(feature_id = row_number()) %>%
    pivot_longer(
      -feature_id,
      names_to = "sample",
      values_to = "signal"
    ) %>%
    mutate(
      log_signal = log2(signal + 1),
      condition = case_when(
        grepl("_mESC_", sample) ~ "mESC",
        grepl("_EpiLC_", sample) ~ "EpiLC",
        grepl("_NE_", sample) ~ "NE"
      )
    ) %>%
    filter(!is.na(condition)) %>%
    mutate(condition = factor(condition, levels = c("mESC", "EpiLC", "NE")))

  stats <- df_long %>%
    pairwise_t_test(
      log_signal ~ condition,
      paired = TRUE,
      p.adjust.method = "BH"
    ) %>%
    add_y_position(fun = "max")

  p <- ggboxplot(
    df_long,
    x = "condition",
    y = "log_signal",
    fill = "condition",
    color = "black",                         # box outlines
    palette = c("#79C9D1", "#FFDC92", "#FF9792"),
    add = "jitter",
    add.params = list(
      size = 0.6,
      alpha = 0.3,
      color = "condition"                   # point colours
    )
  ) +
    scale_color_manual(values = c(
      mESC  = "#4FAAB2",
      EpiLC = "#FFD170",
      NE    = "#FF7770"
    )) +
    stat_pvalue_manual(
      stats,
      label = "p.adj.signif",
      tip.length = 0.01,
      step.increase = 0.1
    ) +
    labs(
      title = paste0("ASCL1 - ", label),
      y = expression(log[2](signal + 1)),
      x = ""
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(size = 10),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 10),
      legend.position = "none"
    ) +
    scale_y_continuous(
      breaks = c(10, 15, 20),
      limits = c(10, 21.5)
    )

  ggsave(
    p,
    filename = file.path(outdir, paste0("Fig2G", label, ".png")),
    bg = "white",
    dpi = 300,
    width = 2,
    height = 3
  )

  list(plot = p, stats = stats)
}

res_neonly    <- plot_ascl1_signal(neonly, "NEonly")
res_neonly$plot

res_notneonly <- plot_ascl1_signal(notneonly, "NotNEonly")
res_notneonly$plot
```


# FIG S2

Fig S2B - correlation matrix of peaksets across replicates / conditions 

```{r, messages=FALSE}

library(here)
library(DiffBind)
library(tidyverse)
library(ggplot2)
library(AnnotationHub)

diffbind_samples <- read.csv(here("../Data/ChIPseq_ASCL1_samples.csv"))

chip_dba <- dba(sampleSheet = diffbind_samples, minOverlap = 1) # this includes all peaks

# built-in blacklist function only uses ENCODE, mm39 not yet available so import custom 
# import from https://github.com/dozmorovlab/excluderanges 
ah <- AnnotationHub()
# mm39 ID is AH107321
chip_blacklist_mm39 <- ah[["AH107321"]]
rm(ah)

# filtering the peakset with the blacklist data 
chip_dba <- dba.blacklist(chip_dba, blacklist = chip_blacklist_mm39, greylist = FALSE)
# Removed: 6677 of 140238 intervals.
# Removed: 2476 merged (of 57267) and 30318 (of 57267) consensus.

chip_consensus_perCond <- dba.peakset(chip_dba, consensus = c(DBA_TISSUE,DBA_CONDITION), minOverlap = 2)
chip_consensus_full <- dba(chip_consensus_perCond, mask = chip_consensus_perCond$masks$Consensus, minOverlap = 1)
chip_consensus_peakset <- dba.peakset(chip_consensus_full, bRetrieve = TRUE, DataType = DBA_DATA_FRAME)

dba.plotHeatmap(chip_consensus_perCond, 
                ColAttributes = DBA_CONDITION,
                colSideCols = c("#79C9D1", "#FDD163", "#FB7D80"),
                colScheme = "Greys")

png(filename = here("../Figures/FigS2B.png"), width = 7, height = 7, units = "in", res = 300)
dba.plotHeatmap(chip_consensus_perCond, 
                ColAttributes = DBA_CONDITION,
                colSideCols = c("#79C9D1", "#FDD163", "#FB7D80"),
                colScheme = "Greys")
dev.off()

```

Fig S2C - PCA plot for ChIP-seq samples with normalisation 

```{r, messages=FALSE}
# pca plots for normalised chip-seq counts 

library(DiffBind)
library(here)

# Load the DBA object
dba <- readRDS(file = here("../Data/ChIPseq_ASCL1_normalised.Rds"))

# Plot PCA 
colours_3 <- c(mESC = "#79C9D1", EpiLC = "#FDD163", NE ="#FB7D80")

dba.plotPCA(dba,
            attributes = DBA_CONDITION,
            score = DBA_SCORE_READS, # DBA_SCORE_READS for library size normalisation, DBA_SCORE_NORMALISED for reads in peaks with RLE
            vColors = unlist(colours_3))

dba.plotPCA(dba,
            attributes = DBA_CONDITION,
            score = DBA_SCORE_NORMALIZED,
            vColors = unlist(colours_3))


png(file = here("../Figures/FigS2C.png"),res = 300, units = "in", width = 5, height = 4)
dba.plotPCA(dba,
            attributes = DBA_CONDITION,
            score = DBA_SCORE_NORMALIZED,
            vColors = unlist(colours_3))
dev.off()

```

Fig S2D - deeptools heatmap across replicates per condition 

```{bash}

computeMatrix scale-regions -S \
../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_1_CPM.bw \
../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_2_CPM.bw \
../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_3_CPM.bw \
../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_4_CPM.bw \
../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
-R ../Data/ChIPseq_ASCL1_mesc10k.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o ../Data/deeptools/matrix/matrix_ascl1chip_mescreps.mat.gz

plotHeatmap -m ../Data/deeptools/matrix/matrix_ascl1chip_mescreps.mat.gz -out ../Data/deeptools/plots/ascl1_mescreps.png --whatToShow 'heatmap and colorbar' --samplesLabel "Rep1" "Rep2" "Rep3" "Rep4" "Avg" --startLabel "" --endLabel "" --colorMap Greys Greys Greys Greys Greys --zMin 0 0 0 0 0 --zMax 100 100 100 100 100 --sortRegions descend

```

Fig S2E - annotation of peaks in mESC, EpiLC and NE to show enhancer / promoter binding sites

```{r, messages=FALSE}

library(here)
library(tidyverse)
# library(ChIPpeakAnno)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm39.knownGene)

# Using the consensus peaks defined and exported from Fig 14 
mesc <- read_table(file = here("../Data/ChIPseq_ASCL1_mesc10k.bed"), col_names = FALSE, show_col_types = F) %>% 
  setNames(c("chr", "start", "end", "peakID")) %>% GRanges()
epilc <- read_table(file = here("../Data/ChIPseq_ASCL1_epilc10k.bed"), col_names = FALSE, show_col_types = F) %>%
  setNames(c("chr", "start", "end", "peakID")) %>% GRanges()
ne <- read_table(file = here("../Data/ChIPseq_ASCL1_ne10k.bed"), col_names = FALSE, show_col_types = F) %>%
  setNames(c("chr", "start", "end", "peakID")) %>% GRanges()

peaks <- list(mESC = mesc, EpiLC = epilc, NE = ne)

annoList <- lapply(peaks, annotatePeak, tssRegion = c(-1000, 1000), TxDb = TxDb.Mmusculus.UCSC.mm39.knownGene, annoDb = "org.Mm.eg.db", addFlankGeneInfo = TRUE, flankDistance = 50000)

annot_colours <- c("Promoter" = "#633372", "5' UTR" = "#9f5691", "3' UTR" = "#aa7aa1", "1st Exon" = "#9b3441", "Other Exon" = "#de597c", "1st Intron" = "#fe9b00", "Other Intron" = "#fbe183", "Downstream (<=300)" = "#f4c40f", "Distal Intergenic" = "#1f6e9c")

p <- plotAnnoBar(annoList, title = "Annotated features of top 10,000 peaks \nper cell type") + scale_fill_manual(values = annot_colours) + theme(title = element_blank())

p

png(file = here("../Figures/FigS2E.png"), res = 300, units = "cm", height = 6, width = 12)
p
dev.off()

```

Fig S2H - correlation heatmap for published ASCL1 ChIP-seq binding sites

```{r, messages=FALSE}

# distance matrix for published chip-seq sets

library(here)
library(tidyverse)
library(DiffBind)
library(pheatmap)
library(MetBrewer)

##### read in our chip data 

mesc <- read_delim((file = here("../Data/ChIPseq_ASCL1_mesc10k.bed")), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

epilc <- read_delim((file = here("../Data/ChIPseq_ASCL1_epilc10k.bed")), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

ne <- read_delim((file = here("../Data/ChIPseq_ASCL1_ne10k.bed")), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)


##### read in the published data 

borromeo2014 <- read_delim(file = here("../Data/chipseq_published/borromeo2014/borromeo2014_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

webb2013 <- read_delim(file = here("../Data/chipseq_published/webb2013/webb2013_adultNPC_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

raposo2015_early <- read_delim(file = here("../Data/chipseq_published/raposo2015/raposo2015_early_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

raposo2015_late <- read_delim(file = here("../Data/chipseq_published/raposo2015/raposo2015_late_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

casey2018 <- read_delim(file = here("../Data/chipseq_published/casey2018/casey2018_mesc_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

wapinski2013_mef <- read_delim(file = here("../Data/chipseq_published/wapinski2013/mm39_wapinski2013_MEF.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

wapinski2013_npc<- read_delim(file = here("../Data/chipseq_published/wapinski2013/mm39_wapinski2013_NPC.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)

wang2023<- read_delim(file = here("../Data/chipseq_published/wang2023/wang2023_mef_mm39.bed"), col_names = FALSE, show_col_types = FALSE) %>%
  dplyr::rename(seqnames = X1, start = X2, end = X3) %>%
  makeGRangesFromDataFrame(keep.extra.columns = FALSE)


# make a list of the peak sets 

peak_sets <- list(
  mESC = mesc,
  EpiLC = epilc,
  NE = ne,
  NeuralTubeE12.5 = borromeo2014,
  AdultNPCs = webb2013,
  NSCs_30min = raposo2015_early,
  NSCs_18hr = raposo2015_late,
  mESCSerumLIF = casey2018,
  MEF_1 = wapinski2013_mef,
  NSCs = wapinski2013_npc,
  MEF_2 = wang2023
)

all_peaks <- Reduce(c, peak_sets)
consensus_peaks <- reduce(all_peaks)  # Merge overlapping regions, there are 109,788 peaks in this set, most seem to come from the wang set 

presence_matrix <- sapply(peak_sets, function(peaks) {
  overlapsAny(consensus_peaks, peaks)
})

# Convert to numeric (TRUE/FALSE → 1/0)
presence_matrix <- apply(presence_matrix, 2, as.integer) # creates a matrix that converts TRUE/FALSE into binary 
rownames(presence_matrix) <- paste0("region_", seq_along(consensus_peaks))

# Distance matrix using Jaccard distance or Euclidean
distance_matrix <- dist(t(presence_matrix), method = "binary")  # Jaccard-style

# OR: Correlation matrix (e.g., Pearson)
cor_matrix <- cor(presence_matrix, method = "pearson")

p <- pheatmap(cor_matrix,
         main = "Correlation Heatmap of Peak Sets",
         clustering_distance_rows = 'euclidean',
         clustering_distance_cols = "euclidean",show_colnames = FALSE,treeheight_row = 0,
         color = met.brewer(name = "VanGogh3", direction = 1)
)

png(filename = here("../Figures/FigS2G.png"), height = 5, width = 6, units = "in", res = 300)
p
dev.off()

```

Fig S2I - plot profiles of ASCL1 binding strength at mESC, NSC, and neural tube binding sites

```{bash}

MATRIX_OUT="../Data/deeptools/matrix/matrix_ascl1chip_caseymESCsites.mat.gz"

computeMatrix scale-regions -S \
  ../Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
  ../Data/chipseq_ascl1-ha/bigwig/gG5_EpiLC_HA_RLE_merged.bigwig \
  ../Data/chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
  -R ../Data/chipseq_published/casey2018/casey2018_mesc_mm39.bed \
  --beforeRegionStartLength 1000 \
  --regionBodyLength 500 \
  --afterRegionStartLength 1000 \
  --missingDataAsZero \
  -o $MATRIX_OUT

plotProfile \
  -m $MATRIX_OUT \
  -out ../Data/deeptools/plots/ESCsites_profile.png \
  --perGroup \
  --samplesLabel mESC EpiLC NE \
  --colors "#79C9D1" "#FFDC92" "#FF9792"
 
computeMatrix scale-regions -S \
bigwigs/gG5_mESC_HA_RLE_merged.bigwig \
bigwigs/gG5_EpiLC_HA_RLE_merged.bigwig \
bigwigs/gG5_NE_HA_RLE_merged.bigwig \
-R beds/mm39_wapinski2013_NPC_501bp.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_ascl1chip_wapinskiNPCsites.mat.gz -p 70
 
plotProfile -m matrix/matrix_ascl1chip_wapinskiNPCsites.mat.gz -out plots/NSCsites_profile.png --perGroup --samplesLabel mESC EpiLC NE --colors "#79C9D1" "#FFDC92" "#FF9792"
 
computeMatrix scale-regions -S \
bigwigs/gG5_mESC_HA_RLE_merged.bigwig \
bigwigs/gG5_EpiLC_HA_RLE_merged.bigwig \
bigwigs/gG5_NE_HA_RLE_merged.bigwig \
-R beds/borromeo2014_mm39.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_ascl1chip_borromeoNTsites.mat.gz -p 70
 
plotProfile -m matrix/matrix_ascl1chip_borromeoNTsites.mat.gz -out plots/NTsites_profile.png --perGroup --samplesLabel mESC EpiLC NE --colors "#79C9D1" "#FFDC92" "#FF9792"

```


# FIG 3

Fig 3B - PCA plot of ATAC-seq samples 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DiffBind)

atac <- readRDS(file = here("../Data/ATACseq_analysed_RLE.Rds"))


colours_9 <- list("mESC_0h" = "#AAE4E9","mESC_6h" = "#79C9D1","mESC_24h" = "#4FAAB2",
                  "EpiLC_0h" = "#FFE8B9","EpiLC_6h" = "#FFDC92","EpiLC_24h" = "#FFD170",
                  "NE_0h" = "#FFBCB9","NE_6h" = "#FF9792","NE_24h" = "#FF7770")

pca1 <- dba.plotPCA(atac, attributes = DBA_TREATMENT, vColors = unlist(colours_9))

png(file = here("../Figures/Fig3B.png"), width = 6, height = 5, units = "in", res = 300)
pca1
dev.off()

```

Fig 3C - deeptools heatmap of ascl1 chip sites split by open or closed in mesc, epilc and ne 

```{bash}

computeMatrix scale-regions -S \
bigwigs/gG5_mESC_HA_RLE_merged.bigwig \
bigwigs/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
-R beds/chip_mescOPEN_t0.bed \
beds/chip_mescCLOSED_t0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_chipatac0_mesc.mat.gz -p 70
 
plotHeatmap -m matrix/matrix_chipatac0_mesc.mat.gz -out plots/chipatac_mesc.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend


computeMatrix scale-regions -S \
bigwigs/gG5_EpiLC_HA_RLE_merged.bigwig \
bigwigs/gG5_EpiLC_0h_ATAC_TMMmerged.bigwig \
-R beds/chip_epilcOPEN_t0.bed \
beds/chip_epilcCLOSED_t0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_chipatac0_epilc.mat.gz -p 70
 
plotHeatmap -m matrix/matrix_chipatac0_epilc.mat.gz -out plots/chipatac_epilc.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend

 
computeMatrix scale-regions -S \
bigwigs/gG5_NE_HA_RLE_merged.bigwig \
bigwigs/gG5_NE_0h_ATAC_TMMmerged.bigwig \
-R beds/chip_neOPEN_t0.bed \
beds/chip_neCLOSED_t0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_chipatac0_ne.mat.gz -p 70
 
plotHeatmap -m matrix/matrix_chipatac0_ne.mat.gz -out plots/chipatac_ne.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend


## global overview of atac and chip 

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_EpiLC_HA_RLE_merged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_mESC_24h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_EpiLC_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_EpiLC_24h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_ASCL1_mesconly10k.bed \
ChIPseq_ASCL1_mescepilc10k.bed \
ChIPseq_ASCL1_epilconly10k.bed \
ChIPseq_ASCL1_epilcne10k.bed \
ChIPseq_ASCL1_neonly10k.bed \
ChIPseq_ASCL1_mescne10k.bed \
ChIPseq_ASCL1_universal10k.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_full_024_allcells.mat.gz

plotHeatmap -m matrix/matrix_full_024_allcells.mat.gz -out plots/ascl1_chip_atac.png --whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ASCL1" "ASCL1" "ATAC no" "ATAC no" "ATAC no" "ATAC 24" "ATAC 24" "ATAC 24" --startLabel "" --endLabel "" --colorMap Greys Greys Greys Greens Greens Greens Greens Greens Greens --zMin 0 0 0 0 0 0 0 0 0 --zMax 100 100 100 25 25 25 25 25 25 --sortRegions descend

```

Fig 3D-E - deeptools heatmap of sites that are bound in NE only or mESC as well 

```{bash}

# run the Data dir for publication on RFS 

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_ne.mat.gz -out deeptools/plots/653_ne.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend


##

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_openT0.bed \
ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne_split.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_ne_split.mat.gz -out deeptools/plots/653_ne_split.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend

##

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_mESC_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_openT0.bed \
ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_mesc_split.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_mesc_split.mat.gz -out deeptools/plots/653_mesc_split.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" "ATAC 24h" --startLabel "" --endLabel "" \
--colorMap Greys Greens Greens --zMin 0 0 0 --zMax 100 25 25 --sortRegions descend

##

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_openT0.bed \
ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne_allT_split.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_ne_allT_split.mat.gz -out deeptools/plots/653_ne_split_t024.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" "ATAC 24h" --startLabel "" --endLabel "" \
--colorMap Greys Greens Greens --zMin 0 0 0 --zMax 100 25 25 --sortRegions descend

##

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_6h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_mESC_6h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_mESC_24h_ATAC_TMMmerged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_EpiLC_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_EpiLC_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_EpiLC_6h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_EpiLC_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_all_allT.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_all_allT.mat.gz -out deeptools/plots/653_all_allT.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" "ATAC 6h" "ATAC 24h" "ASCL1" "ATAC 0h" "ATAC 6h" "ATAC 24h" "ASCL1" "ATAC 0h" "ATAC 6h" "ATAC 24h" --startLabel "" --endLabel "" \
--colorMap Greys Greens Greens Greens --zMin 0 0 0 0 0 0 0 0 0 0 0 0 --zMax 100 25 25 25 100 25 25 25 100 25 25 25 --sortRegions descend

```

Fig 3F - donut plots of percentage of sites that are bound & accessible in mESC and NE 

```{r, messages=FALSE}

library(here)
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(patchwork)

all_653 <- read.table(file = here("../Data/ChIPseq_653NEsites.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
all_653.gr <- all_653 %>% GRanges()

# 417 are not in mESC
not_mesc <- read.table(file = here("../Data/ChIPseq_653NEsites_splitnotmesc.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
not_mesc.gr <- not_mesc %>% GRanges()

# 236 are in mESC
also_mesc <- read.table(file = here("../Data/ChIPseq_653NEsites_splitmesc.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
also_mesc.gr <- also_mesc %>% GRanges()


### 653 ne sites 
all_653 <- read.table(file = here("../Data/ChIPseq_653NEsites.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
all_653.gr <- all_653 %>% GRanges()

## mesc atac t0
mesc.atac <- read.table(file = here("../Data/ATACseq_mESC_0h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
mesc.atac.gr <- mesc.atac %>% GRanges()

# Overlapping (min 50bp overlap)
all_362_mesc_atac <- subsetByOverlaps(all_653.gr, mesc.atac.gr, minoverlap = 50)

# mESC_NE and Universal
mesc_universal.gr <- all_362_mesc_atac[all_362_mesc_atac$peakset %in% c("mESC_NE", "Universal")]

# NE and EpiLC_NE
ne_epilc.gr <- all_362_mesc_atac[all_362_mesc_atac$peakset %in% c("NE", "EpiLC_NE")]

cat("mESC_NE + Universal:", length(mesc_universal.gr), "\n")
cat("NE + EpiLC_NE:", length(ne_epilc.gr), "\n")

# mESC_NE and Universal
write.table(as.data.frame(mesc_universal.gr), 
            file = here("../Data/ChIPseq_362_openmesc_boundmesc.bed"), 
            sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# NE and EpiLC_NE
write.table(as.data.frame(ne_epilc.gr), 
            file = here("../Data/ChIPseq_362_openmesc_notboundmesc.bed"), 
            sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# do these sites overlap with atac 0hr peaks ?

# load in the mesc peaks t=0 
mesc.atac <- read.table(file = here("../Data/ATACseq_mESC_0h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
mesc.atac.gr <- mesc.atac %>% GRanges()

# load in the mesc peaks t=24
mesc.atac.24 <- read.table(file = here("../Data/ATACseq_mESC_24h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
mesc.atac.24.gr <- mesc.atac.24 %>% GRanges()

# load in the ne peaks t=0
ne.atac <- read.table(file = here("../Data/ATACseq_NE_0h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
ne.atac.gr <- ne.atac %>% GRanges()


# find overlaps 

calc_open <- function(gr_sites, gr_atac) {

  hits <- countOverlaps(gr_sites, gr_atac, minoverlap = 50) > 0

  data.frame(
    status = c("Open", "Closed"),
    count = c(sum(hits), sum(!hits))
  )
}

not_mesc_mescATAC <- calc_open(not_mesc.gr, mesc.atac.gr)
not_mesc_neATAC   <- calc_open(not_mesc.gr, ne.atac.gr)

also_mesc_mescATAC <- calc_open(also_mesc.gr, mesc.atac.gr)
also_mesc_neATAC   <- calc_open(also_mesc.gr, ne.atac.gr)

all653_ne <- calc_open(all_653.gr, ne.atac.gr)

all653_mesc <- calc_open(all_653.gr, mesc.atac.gr)

all653_mesc24 <- calc_open(all_653.gr, mesc.atac.24.gr)

# plot donut plot to show % of sites that are open (e.g overlap with atac peak) in mESC and then in NE, for each of the two sets of ChIPseq sites - so four plots in total

plot_donut <- function(df) {

  df <- df %>%
    mutate(status = factor(status, levels = c("Closed", "Open"))) %>%
    arrange(status)

  pct_open <- round(100 * df$count[df$status == "Open"] / sum(df$count), 1)

  ggplot(df, aes(x = 2, y = count, fill = status)) +
    geom_col(width = 1, colour = NA) +
    coord_polar(theta = "y", direction = -1) + 
    xlim(0.5, 2.5) + # this makes it donut
    scale_fill_manual(values = c("grey", "#00cc00")) +
    theme_void() +
    theme(legend.position = "none")
}

p0 <- plot_donut(all653_ne)
ggsave(p0, filename = here("../Figures/Fig4F-ne0.png"), bg = "white", dpi = 300, height = 3, width = 3)

p653_mesc0 <- plot_donut(all653_mesc)
ggsave(filename = here("../Figures/Fig4F-mesc0.png"), bg = "white", dpi = 300, height = 3, width = 3)

p653_mesc24 <- plot_donut(all653_mesc24)
ggsave(filename = here("../Figures/Fig4F-mesc24.png"), bg = "white", dpi = 300, height = 3, width = 3)

p1 <- plot_donut(not_mesc_mescATAC)
ggsave(filename = here("../Figures/Fig4E-1.png"), bg = "white", dpi = 300, height = 3, width = 3)

p2 <- plot_donut(not_mesc_neATAC)
ggsave(filename = here("../Figures/Fig4E-2.png"), bg = "white", dpi = 300, height = 3, width = 3)

p3 <- plot_donut(also_mesc_mescATAC)
ggsave(filename = here("../Figures/Fig4F-1.png"), bg = "white", dpi = 300, height = 3, width = 3)

p4 <- plot_donut(also_mesc_neATAC)
ggsave(filename = here("../Figures/Fig4F-2.png"), bg = "white", dpi = 300, height = 3, width = 3)

```

```{r, messages=FALSE}

library(here)
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(patchwork)

# 653 are not in mESC
nesites <- read.table(file = here("../Data/ChIPseq_653NEsites.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
nesites.gr <- nesites %>% GRanges()

# do these sites overlap with atac 0hr peaks ?

# load in the ne peaks t=0
ne.atac0 <- read.table(file = here("../Data/ATACseq_NE_0h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
ne.atac0.gr <- ne.atac0 %>% GRanges()

# load in the ne peaks t=24
ne.atac24 <- read.table(file = here("../Data/ATACseq_NE_24h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
ne.atac24.gr <- ne.atac24 %>% GRanges()


calc_open <- function(gr_sites, gr_atac) {

  hits <- countOverlaps(gr_sites, gr_atac, minoverlap = 50) > 0

  data.frame(
    status = c("Open", "Closed"),
    count = c(sum(hits), sum(!hits))
  )
}

nesites.open0 <- calc_open(nesites.gr, ne.atac0.gr)
nesites.open24 <- calc_open(nesites.gr, ne.atac24.gr)

nesites.open0 <- nesites.gr[countOverlaps(nesites.gr, ne.atac0.gr, minoverlap = 50) > 0]

nesites.closed0 <- nesites.gr[!countOverlaps(nesites.gr, ne.atac0.gr, minoverlap = 50) > 0]

nesites.open0.export <- nesites.open0 %>% as.data.frame()
nesites.closed0.export <- nesites.closed0 %>% as.data.frame()

# 160 NE sites are closed at t 0
write.table(nesites.closed0.export, file = here("../Data/ChIPseq_653NEsites_closedT0.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

# 493 NE sites are open at t 0
write.table(nesites.open0.export, file = here("../Data/ChIPseq_653NEsites_openT0.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

pNEt24 <- plot_donut(nesites.open24)
ggsave(pNEt24, filename = here("../Figures/Fig4F-ne24.png"), bg = "white", dpi = 300, height = 3, width = 3)

```

```{bash}

computeMatrix scale-regions -S \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne_t024.mat.gz

plotProfile -m deeptools/matrix/matrix_653_ne_t024.mat.gz \
-out deeptools/plots/closedNE_atac_profile.png --perGroup --samplesLabel 0 24 \
--colors "#00cc00" "#008000" --startLabel "" --endLabel "" --yMax 10

computeMatrix scale-regions -S \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_openT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne_t024_open.mat.gz

plotProfile -m deeptools/matrix/matrix_653_ne_t024_open.mat.gz \
-out deeptools/plots/openNE_atac_profile.png --perGroup --samplesLabel 0 24 \
--colors "#00cc00" "#008000" --startLabel "" --endLabel "" --yMax 30 --plotHeight 6 --plotWidth 6

####

computeMatrix scale-regions -S \
atacseq/bigwig/gG5_NE_0h_ATAC_TMMmerged.bigwig \
atacseq/bigwig/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites_openT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_ne_t024_open.mat.gz

plotProfile -m deeptools/matrix/matrix_653_ne_t024_open.mat.gz \
-out deeptools/plots/openNE_atac_profile.png --perGroup --samplesLabel 0 24 \
--colors "#00cc00" "#008000" --startLabel "" --endLabel "" --yMax 30 --plotHeight 6 --plotWidth 6


####

computeMatrix scale-regions -S \
bigwigs/gG5_NE_0h_ATAC_TMMmerged.bigwig \
bigwigs/gG5_NE_6h_ATAC_TMMmerged.bigwig \
bigwigs/gG5_NE_24h_ATAC_TMMmerged.bigwig \
-R beds/ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_ascl1chip_closedNE_NE.mat.gz -p 70
 
plotProfile -m matrix/matrix_ascl1chip_closedNE_NE.mat.gz \
-out plots/closedNE_neatac_profile.png --perGroup --samplesLabel 0 6 24 \
--colors "#00e600" "#00cc00" "#008000" --startLabel "" --endLabel "" --yMax 10


computeMatrix scale-regions -S \
bigwigs/gG5_EpiLC_0h_ATAC_TMMmerged.bigwig \
bigwigs/gG5_EpiLC_6h_ATAC_TMMmerged.bigwig \
bigwigs/gG5_EpiLC_24h_ATAC_TMMmerged.bigwig \
-R beds/ChIPseq_653NEsites_closedT0.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o matrix/matrix_ascl1chip_closedNE_EpiLC.mat.gz -p 70
 
plotProfile -m matrix/matrix_ascl1chip_closedNE_EpiLC.mat.gz \
-out plots/closedNE_epilcatac_profile.png --perGroup --samplesLabel 0 6 24 \
--colors "#00e600" "#00cc00" "#008000" --startLabel "" --endLabel "" --yMax 10

```

```{r, messages=FALSE}

library(GenomicRanges)
library(here)
library(tidyverse)

# 653 are not in mESC
nesites <- read.table(file = here("../Data/ChIPseq_653NEsites.bed"), sep = "\t", col.names = c("seqnames", "start", "end", "width", "strand", "peakID", "peakset"))
nesites.gr <- nesites %>% GRanges()

# do these sites overlap with atac 0hr peaks ?

# load in the ne peaks t=0
ne.atac0 <- read.table(file = here("../Data/ATACseq_NE_0h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
ne.atac0.gr <- ne.atac0 %>% GRanges()

# load in the ne peaks t=24
ne.atac24 <- read.table(file = here("../Data/ATACseq_NE_24h_peaks_in3of4.bed"), col.names = c("seqnames", "start", "end", "peakID"))
ne.atac24.gr <- ne.atac24 %>% GRanges()

nesites.closed0 <- nesites.gr[!countOverlaps(nesites.gr, ne.atac0.gr, minoverlap = 50) > 0] # 160 of these are closed 

nesites.opened <- nesites.closed0[countOverlaps(nesites.closed0, ne.atac24.gr, minoverlap = 50 ) > 0]

nesites.opened.export <- nesites.opened %>% as.data.frame()

# 109/160 NE sites are closed at t 0 and open at t24 
write.table(nesites.opened.export, file = here("../Data/ChIPseq_653NEsites_opened.bed"), sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)

```


# FIG S3

Fig S3A - number of ATAC-seq peaks in each sample 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DiffBind)
library(AnnotationHub)
library(cowplot)

atac_samples <- read.csv(here("../Data/ATACseq/atacseq_diffbindSamples.csv"))

# built-in blacklist function only uses ENCODE, mm39 not yet available so import custom 
# import from https://github.com/dozmorovlab/excluderanges 
ah <- AnnotationHub()
# mm39 ID is AH107321
atac_blacklist_mm39 <- ah[["AH107321"]]
rm(ah)

atac_data <- dba(sampleSheet = atac_samples, minOverlap = 1) # this includes all peaks
atac_data <- dba.blacklist(atac_data, blacklist = atac_blacklist_mm39, greylist = FALSE) 
# Applying blacklist...
# Removed: 198227 of 4703301 intervals.
# Removed: 13827 merged (of 355316) and 81704 (of 355316) consensus.

atac_consensus_perCond <- dba.peakset(atac_data, consensus = DBA_TREATMENT, minOverlap = 3) # this takes peaks that are in 3/4 samples 
atac_consensus_full <- dba(atac_consensus_perCond, mask = atac_consensus_perCond$masks$Consensus, minOverlap = 1)



peaks_df <- dba.show(atac_data) %>% as.data.frame()
peaks_df$Treatment <- factor(peaks_df$Treatment, levels = c("mESC_0h", "mESC_6h", "mESC_24h", "EpiLC_0h", "EpiLC_6h", "EpiLC_24h", "NE_0h", "NE_6h", "NE_24h"))

cons_df <- dba.show(atac_consensus_full) %>% as.data.frame()
cons_df$Treatment <- factor(cons_df$Treatment, levels = c("mESC_0h", "mESC_6h", "mESC_24h", "EpiLC_0h", "EpiLC_6h", "EpiLC_24h", "NE_0h", "NE_6h", "NE_24h"))


colours_9 <- list("mESC_0h" = "#AAE4E9","mESC_6h" = "#79C9D1","mESC_24h" = "#4FAAB2",
                  "EpiLC_0h" = "#FFE8B9","EpiLC_6h" = "#FFDC92","EpiLC_24h" = "#FFD170",
                  "NE_0h" = "#FFBCB9","NE_6h" = "#FF9792","NE_24h" = "#FF7770")

peakplot <- ggplot(data = subset(peaks_df, grepl("_", ID)), aes(x = Treatment, y = Intervals/1000, fill = Treatment)) +
  geom_boxplot(aes(fill = Treatment), outlier.shape = NA) +  # Boxplot retains the legend
  geom_point(position=position_jitterdodge(), size = 4, shape = 16, color = "black") +
  geom_point(data = cons_df, aes(x = Treatment, y = Intervals/1000, color = "Consensus set \n in 3 of 4 \n replicates"), size = 3, shape = 8) +  # Add legend for consensus points
  scale_fill_manual(name = "", values = colours_9) + 
  scale_color_manual(name = "", values = c("Consensus set \n in 3 of 4 \n replicates" = "black")) +  # Custom legend entry for consensus points
  guides(fill = guide_legend(override.aes = list(shape = NA))) + # this excludes the shape being plotted for the boxplot legend 
  theme_cowplot() +
  theme(legend.position = "none",  # Adjust as needed
        strip.text.x = element_text(size = 10, face = "bold"),
        axis.title = element_text(size = 20),  # Axis titles font size 20
        axis.text = element_text(size = 16),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  labs(title = element_blank(),
       x = element_blank(),
       y = expression(paste("Number of peaks (", 10^3, ")")))

peakplot

ggsave(plot = peakplot, filename = here("../Figures/FigS3A.png"), dpi = 300, width = 5, height = 5, units = "in", bg= "white")

```

Fig S3B - FRiP % of ATAC-seq samples

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DiffBind)
library(cowplot)

atac_diffbind <- readRDS(file = here("../Data/ATACseq_counted_nochrUn.Rds"))

info <- dba.show(atac_diffbind)
info$Treatment <- factor(info$Treatment, levels = c("mESC_0h", "mESC_6h", "mESC_24h", "EpiLC_0h", "EpiLC_6h", "EpiLC_24h", "NE_0h", "NE_6h", "NE_24h"))
info <- info %>% mutate("frip_pcnt" = FRiP * 100)

colours_9 <- list("mESC_0h" = "#AAE4E9","mESC_6h" = "#79C9D1","mESC_24h" = "#4FAAB2",
                  "EpiLC_0h" = "#FFE8B9","EpiLC_6h" = "#FFDC92","EpiLC_24h" = "#FFD170",
                  "NE_0h" = "#FFBCB9","NE_6h" = "#FF9792","NE_24h" = "#FF7770")

frip <- ggplot(info, aes(x = Treatment, y = frip_pcnt, fill = Treatment)) +
  geom_boxplot(show.legend = FALSE) +
  # geom_bar(stat = "summary", fun = mean) +
  scale_fill_manual(values = colours_9) +
  geom_point(position=position_jitterdodge(), size = 4, shape = 16, color = "black") +
  labs(title = NULL,
       x = NULL,
       y = "FRiP (%)"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 40)) +
  theme_cowplot() +
  theme(
    axis.title = element_text(size = 20),  # Axis titles font size 20
    axis.text = element_text(size = 16),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none"
  )

frip

ggsave(plot = frip, filename = here("../Figures/FigS3B.png"), width = 5, height = 5, device = "png", dpi = 300, bg= "white")

```

Fig S3C - TSS enrichment 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(ggplot2)
library(cowplot)

tss_reads <- read_delim(file = here("../Data/ATACseq/tss_enrichment/readCountInPeaks.txt.summary"), show_col_types = F)
flank_reads <- read_delim(file = here("../Data/ATACseq/tss_enrichment/readCountInFlanks.txt.summary"), show_col_types = F)

# Convert to long format and extract sample names
tss_reads_long <- tss_reads %>%
  pivot_longer(-Status, names_to = "Sample", values_to = "TSS Reads") %>%
  mutate(Sample = basename(Sample) %>% str_remove(".nodup.bam")) %>%
  filter(Status == "Assigned")  # Keep only Assigned reads

flank_reads_long <- flank_reads %>%
  pivot_longer(-Status, names_to = "Sample", values_to = "Flank Reads") %>%
  mutate(Sample = basename(Sample) %>% str_remove(".nodup.bam")) %>%
  filter(Status == "Assigned")  # Keep only Assigned reads

summary <- tss_reads_long %>% right_join(flank_reads_long) %>% mutate("Enrichment" = `TSS Reads`/`Flank Reads`) %>% mutate("Cell" = str_split_i(Sample, "_", i = 2)) %>% mutate("Time" = str_split_i(Sample, "_", i = 3)) %>% mutate("Condition" = paste(Cell, Time, sep = "_"))
summary$Condition <- factor(summary$Condition, levels = c("mESC_0h", "mESC_6h", "mESC_24h", "EpiLC_0h", "EpiLC_6h", "EpiLC_24h", "NE_0h", "NE_6h", "NE_24h"))

# summary.0h <- summary %>% dplyr::filter(Condition %in% c("mESC_0h", "EpiLC_0h", "NE_0h"))
  
  
# # Compute total Unassigned for each sample
# df_summary <- df_long %>%
#   group_by(Sample) %>%
#   summarise(Assigned = sum(Reads[Status == "Assigned"]),
#             Unassigned = sum(Reads[Status != "Assigned"]))

colours_9 <- list("mESC_0h" = "#AAE4E9","mESC_6h" = "#79C9D1","mESC_24h" = "#4FAAB2",
                  "EpiLC_0h" = "#FFE8B9","EpiLC_6h" = "#FFDC92","EpiLC_24h" = "#FFD170",
                  "NE_0h" = "#FFBCB9","NE_6h" = "#FF9792","NE_24h" = "#FF7770")


# Plot the data
tss <- ggplot(data = summary, aes(x = Condition, y = Enrichment, fill = Condition)) +
  geom_boxplot() +
  geom_point(position=position_jitterdodge(), size = 4, shape = 16, color = "black") +
  scale_fill_manual(values = colours_9) +
  theme_cowplot() +
  labs(title = element_blank(),
       x = element_blank(),
       y = "TSS Enrichment") +
  theme(
    axis.title = element_text(size = 20),  # Axis titles font size 20
    axis.text = element_text(size = 16),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 30))


ggsave(plot = tss, filename = here("../Figures/FigS3C.png"), width = 5, height = 5, device = "png", dpi = 300, bg= "white")

```

Fig S3D - chrM reads

```{r, messages=FALSE}
library(here)
library(tidyverse)

mitochondria <- read_delim(file = here("../Data/ATACseq/mtDNA_content.txt"), show_col_types = FALSE)

mitochondria <- mitochondria %>%
  separate(Sample, into = c("Condition", "Replicate"), sep = "_rep") %>%  # Split Sample into Condition and Replicate
  mutate(
    Replicate = paste0("rep", Replicate),  # Add "rep" back to the Replicate column
    Condition = str_remove(Condition, "^gG5_"),  # Remove the 'gG5_' prefix from Condition
    mtDNA_Content = parse_number(`mtDNA_Content(%)`)  # Convert mtDNA_Content(%) to numeric by removing the %
  )

mitochondria$Condition <- factor(mitochondria$Condition, levels = c("mESC_0h", "mESC_6h", "mESC_24h", "EpiLC_0h", "EpiLC_6h", "EpiLC_24h", "NE_0h", "NE_6h", "NE_24h"))

colours_9 <- list("mESC_0h" = "#AAE4E9","mESC_6h" = "#79C9D1","mESC_24h" = "#4FAAB2",
                  "EpiLC_0h" = "#FFE8B9","EpiLC_6h" = "#FFDC92","EpiLC_24h" = "#FFD170",
                  "NE_0h" = "#FFBCB9","NE_6h" = "#FF9792","NE_24h" = "#FF7770")

pcntMit <- ggplot(mitochondria, aes(x = Condition, y = mtDNA_Content, fill = Condition)) +
  # geom_bar(stat = "summary", fun = mean) +
  geom_boxplot() +
  scale_fill_manual(values = colours_9) +
  geom_point(position=position_jitterdodge(), size = 4, shape = 16, color = "black") +
  labs(title = element_blank(),
       x = element_blank(),
       y = "% Mitochondrial Reads"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5)) +
  theme_cowplot() +
  theme(
    axis.title = element_text(size = 20),  # Axis titles font size 20
    axis.text = element_text(size = 16),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none"
  )

ggsave(plot = pcntMit, filename = here("../Figures/FigS3D.png"), width = 5, height = 5, device = "png", dpi = 300, bg= "white")
```


# FIG 4

Fig 4A - heatmap to show 653 NE sites open or closed in mESC

```{bash}

computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
atacseq/bigwig/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
-R ChIPseq_653NEsites.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_653_mesc.mat.gz

plotHeatmap -m deeptools/matrix/matrix_653_mesc.mat.gz -out deeptools/plots/653_mesc.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" --startLabel "" --endLabel "" \
--colorMap Greys Greens --zMin 0 0 --zMax 100 25 --sortRegions descend

```

Fig 4C-D - mesc open bound and not bound plus histone profile 

```{bash}
##

computeMatrix scale-regions -S \
Data/chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
Data/atacseq/bigwig/gG5_mESC_0h_ATAC_TMMmerged.bigwig \
Data/chipseq_published/cruzmolina2017/2i_H3K27ac_mm39.bw \
Data/chipseq_published/cruzmolina2017/AntNPC_H3K27ac_mm39.bw \
-R Data/ChIPseq_362_openmesc_boundmesc.bed \
Data/ChIPseq_362_openmesc_notboundmesc.bed \
--beforeRegionStartLength 2000 \
--regionBodyLength 500 \
--afterRegionStartLength 2000 \
--missingDataAsZero -o Data/deeptools/matrix/matrix_362_split_mescboundornot.mat.gz

plotHeatmap -m Data/deeptools/matrix/matrix_362_split_mescboundornot.mat.gz -out Data/deeptools/plots/362_mescsplit.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "ASCL1" "ATAC 0h" "H3K27ac 2i" "H3K27ac AntNPC" --startLabel "" --endLabel "" \
--colorMap Greys Greens Blues Blues --zMin 0 0 0 0 --zMax 100 25 10 10 --sortRegions descend


###

computeMatrix scale-regions -S \
Data/chipseq_published/cruzmolina2017/2i_H3K27ac_mm39.bw \
Data/chipseq_published/cruzmolina2017/AntNPC_H3K27ac_mm39.bw \
-R Data/ChIPseq_362_openmesc_boundmesc.bed \
Data/ChIPseq_362_openmesc_notboundmesc.bed \
--beforeRegionStartLength 5000 \
--regionBodyLength 500 \
--afterRegionStartLength 5000 \
--missingDataAsZero -o Data/deeptools/matrix/matrix_362_split_mescboundornot_h3k27.mat.gz

plotProfile -m Data/deeptools/matrix/matrix_362_split_mescboundornot_h3k27.mat.gz -out Data/deeptools/plots/362_mescsplit_h3k27ac.png \
--whatToShow 'heatmap and colorbar' --samplesLabel "H3K27ac 2i" "H3K27ac AntNPC" --startLabel "" --endLabel "" \
--colorMap  Blues --zMin 0 0 --zMax 20 20 --sortRegions descend --plotHeight 10 --plotWidth 10



```

Fig 4E - specific gene counts for Ank2 Cyth1 Hes5 (no EpiLC)

```{r}
library(here)
library(AnnotationHub)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(cowplot)

# load in the rna-seq rds object
# these counts are corrected for library size and Tx length
rna <- readRDS(file = here("../Data/RNAseq_ASCL1_dds.Rds"))
rna <- estimateSizeFactors(rna)
rnaseq_counts <- counts(object = rna, normalized = TRUE)
# 
# # write to a csv 
# write.csv(rnaseq_counts, file = here("../results/countsmatrix_normalised.csv"))

# need to annotate the row names to gene symbols
ah <- AnnotationHub()
# ah object ID for the mm39 108 ensembl build is AH109367
EnsDb.Mmusculus.v108 <- ah[["AH109367"]]
genes <- genes(EnsDb.Mmusculus.v108, return.type = "DataFrame")
gene2symbol <- genes[c(1,12)]
gene2symbol_matched <- gene2symbol$symbol[match(rownames(rnaseq_counts), gene2symbol$gene_id)]
# annotate the rna-seq table
rownames(rnaseq_counts) <- gene2symbol_matched

colours <- c("C11_mESC_t24" = "lightgrey", "C11_NE_t24" = "lightgrey",
             "gG5_mESC_t0" = "#AAE4E9","gG5_mESC_t6" = "#79C9D1","gG5_mESC_t24" = "#4FAAB2",
             "gG5_NE_t0" = "#FFBCB9","gG5_NE_t6" = "#FF9792","gG5_NE_t24" = "#FF7770")


# before generating the heatmap, want to change the order so that the C11 columns are next to the t0 columns 
desired_order <- c(
  # mESC
  "exp1_C11_mESC_t24_rep1", "exp1_C11_mESC_t24_rep2", "exp1_C11_mESC_t24_rep3",
  "exp2_gG5_mESC_t0_rep1", "exp2_gG5_mESC_t0_rep2", "exp2_gG5_mESC_t0_rep3", "exp2_gG5_mESC_t0_rep4",
  "exp2_gG5_mESC_t6_rep1", "exp2_gG5_mESC_t6_rep2", "exp2_gG5_mESC_t6_rep3", "exp2_gG5_mESC_t6_rep4",
  "exp1_gG5_mESC_t24_rep1", "exp1_gG5_mESC_t24_rep2", "exp1_gG5_mESC_t24_rep3", 
  "exp2_gG5_mESC_t24_rep1", "exp2_gG5_mESC_t24_rep2", "exp2_gG5_mESC_t24_rep3", "exp2_gG5_mESC_t24_rep4",
  
  # NE
  "exp1_C11_NE_t24_rep1", "exp1_C11_NE_t24_rep2", "exp1_C11_NE_t24_rep3", 
  "exp2_gG5_NE_t0_rep1", "exp2_gG5_NE_t0_rep2", "exp2_gG5_NE_t0_rep3", "exp2_gG5_NE_t0_rep4",
  "exp2_gG5_NE_t6_rep1", "exp2_gG5_NE_t6_rep2", "exp2_gG5_NE_t6_rep3", "exp2_gG5_NE_t6_rep4",
  "exp1_gG5_NE_t24_rep1", "exp1_gG5_NE_t24_rep2", "exp1_gG5_NE_t24_rep3", 
  "exp2_gG5_NE_t24_rep1", "exp2_gG5_NE_t24_rep2", "exp2_gG5_NE_t24_rep3", "exp2_gG5_NE_t24_rep4"
)

# Reorder the heatmap data and annotation
rnaseq_counts <- rnaseq_counts[, desired_order]

genes.ofi <- c("Ank2", "Cyth1", "Hes5")# <- replace with your list

# Loop through each gene and create + save the plot
for (gene.ofi in genes.ofi) {
  
  # Get counts for current gene
  log.counts.ofi <- log2(rnaseq_counts[gene.ofi, ] + 1)
  counts.ofi <- rnaseq_counts[gene.ofi, ]
  
  # Create plot
  gg.logtpmBox <- counts.ofi %>%
    tibble::as_tibble() %>%
    mutate(sample = names(counts.ofi)) %>%
    separate(sample, c('exp', 'line', 'cell', 'time', 'rep'), sep = '_') %>%
    mutate(cond_time = paste(line, cell, time, sep = "_")) %>%
    mutate(cond_time = factor(cond_time, levels = unique(cond_time))) %>%
    ggplot(aes(x = cond_time, y = value, fill = cond_time)) +
    geom_boxplot(width = 0.7, outlier.shape = NA, color = "black", size = 0.3) +
    geom_point(
      aes(fill = cond_time),
      position = position_jitter(width = 0.1, height = 0),
      size = 2, shape = 21, color = "black"
    ) +
    xlab('') +
    #ylab(bquote(log[2]~"(scaled counts + 1)")) +
    ylab("scaled counts") +
    scale_fill_manual(values = colours) +
    scale_color_manual(values = colours) +
    scale_y_continuous(limits = c(NA, NA)) +
    theme_cowplot() +
    labs(title = gene.ofi) +
    theme(
      legend.position = 'none',
      axis.line.x = element_line(linewidth = 0.4),
      axis.line.y = element_line(linewidth = 0.4),
      axis.line = element_line(linetype = 'solid'),
      axis.text.x = element_blank(),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 16, face = "italic", hjust = 0.5)
    )
  
  # Save the plot
  ggsave(
    plot = gg.logtpmBox,
    filename = here("../Figures/", paste0("RNAseq_", gene.ofi, "_noEpiLC.png")),
    device = "png", units = "cm", width = 5.5, height = 8, bg = "white"
  )
}

```


# FIG S4

Fig S4A - heatmap for ASCL1-ChIPseq peaks split by cell type 

```{bash}
computeMatrix scale-regions -S \
chipseq_ascl1-ha/bigwig/gG5_mESC_HA_RLE_merged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_EpiLC_HA_RLE_merged.bigwig \
chipseq_ascl1-ha/bigwig/gG5_NE_HA_RLE_merged.bigwig \
-R ChIPseq_ASCL1_mesconly10k.bed \
ChIPseq_ASCL1_mescepilc10k.bed \
ChIPseq_ASCL1_epilconly10k.bed \
ChIPseq_ASCL1_epilcne10k.bed \
ChIPseq_ASCL1_neonly10k.bed \
ChIPseq_ASCL1_mescne10k.bed \
ChIPseq_ASCL1_universal10k.bed \
--beforeRegionStartLength 1000 \
--regionBodyLength 500 \
--afterRegionStartLength 1000 \
--missingDataAsZero -o deeptools/matrix/matrix_ascl1chip.mat.gz
```

Fig S4B - heatmap for H3K27ac 

```{bash}

computeMatrix scale-regions -S \
Data/chipseq_published/cruzmolina2017/2i_H3K27ac_mm39.bw \
Data/chipseq_published/cruzmolina2017/AntNPC_H3K27ac_mm39.bw \
-R ChIPseq_ASCL1_mesconly10k.bed \
ChIPseq_ASCL1_mescepilc10k.bed \
ChIPseq_ASCL1_epilconly10k.bed \
ChIPseq_ASCL1_epilcne10k.bed \
ChIPseq_ASCL1_neonly10k.bed \
ChIPseq_ASCL1_mescne10k.bed \
ChIPseq_ASCL1_universal10k.bed \
--beforeRegionStartLength 5000 \
--regionBodyLength 500 \
--afterRegionStartLength 5000 \
--missingDataAsZero -o Data/deeptools/matrix/matrix_h3k27ac_allsites.mat.gz

```

Fig S4C - heatmap for H3K4me1

```{bash}
Data/chipseq_published/bleckwehl2020/GSM4694551_H3K4me1_mESC_mm39.bw \
Data/chipseq_published/cruzmolina2017/GSM4694551_H3K4me1_EpiLC_mm39.bw \
-R ChIPseq_ASCL1_mesconly10k.bed \
ChIPseq_ASCL1_mescepilc10k.bed \
ChIPseq_ASCL1_epilconly10k.bed \
ChIPseq_ASCL1_epilcne10k.bed \
ChIPseq_ASCL1_neonly10k.bed \
ChIPseq_ASCL1_mescne10k.bed \
ChIPseq_ASCL1_universal10k.bed \
--beforeRegionStartLength 5000 \
--regionBodyLength 500 \
--afterRegionStartLength 5000 \
--missingDataAsZero -o Data/deeptools/matrix/matrix_h3k4me1_allsites.mat.gz
```



# FIG 5

Fig 5C - PCA plot for HDACi RNA-seq 

```{r, messages=FALSE}

library(here)
library(tidyverse)
library(DESeq2)
library(cowplot)


# plotting a pca plot for the HDACi RNA-seq experiment 

hdacirna <- readRDS(file = here("../Data/RNAseq_HDACi_dds.rds"))

vsd <- vst(hdacirna, blind = FALSE)

colours <- list(
  "mESC_None_DMSO" = "#AAE4E9", "mESC_Dex_DMSO" = "#4FAAB2", "mESC_None_TSA" = "#AAE4E9", "mESC_Dex_TSA" = "#4FAAB2", 
  "NE_None_DMSO" = "#FFBCB9", "NE_Dex_DMSO" = "#FF7770", "NE_None_TSA" = "#FFBCB9", "NE_Dex_TSA" = "#FF7770")


pcaData.12 <- plotPCA(vsd, intgroup="Group", returnData=TRUE, ntop = 1000)
percentVar.12 <- round(100 * attr(pcaData.12, "percentVar"))
pcaplot.12 <- ggplot(pcaData.12, aes(PC1, PC2, color=Group, shape=TSA)) +
  geom_point(size=4, show.legend = FALSE) +
  xlab(paste0("PC1: ",percentVar.12[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar.12[2],"% variance")) + 
  scale_colour_manual(values = unlist(colours)) + 
  #coord_fixed() + 
  theme_cowplot()

ggsave(pcaplot.12, filename = here("../Figures/Fig5B.png"), device = "png", dpi = 300, height = 8, width = 8, units = "cm", bg = "white")


# plot pc2 pc3 

pcaData.23 <- plotPCA(vsd, intgroup="Group", returnData=TRUE, ntop = 1000, pcsToUse = 2:3)
percentVar.23 <- round(100 * attr(pcaData.23, "percentVar"))

pcaplot.23 <- ggplot(pcaData.23, aes(PC2, PC3, color=Group, shape=TSA)) +
  geom_point(size=4, show.legend = FALSE) +
  xlab(paste0("PC2: ",percentVar.23[1],"% variance")) +
  ylab(paste0("PC3: ",percentVar.23[2],"% variance")) + 
  scale_colour_manual(values = unlist(colours)) + 
  #coord_fixed() + 
  theme_cowplot()

ggsave(pcaplot.23, filename = here("../Figures/FigS5B.png"), device = "png", dpi = 300, height = 8, width = 8, units = "cm", bg = "white")
```

Fig 5D - counts for selected genes in mESC with HDACi

```{r, messages=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(cowplot)

counts <- read.csv(here("../Data/RNAseq_HDACi_normcounts.csv"))
counts <- counts %>% dplyr::rename(gene_id = X)

# Map Ensembl IDs to gene symbols
gene_map <- read.csv(here("../Data/RNAseq_GeneMap.csv"), row.names = 1)
counts <- counts %>%
  left_join(gene_map %>% dplyr::select(gene_id, symbol), by = "gene_id") %>%
  dplyr::filter(!is.na(symbol) & symbol != "") %>%
  dplyr::select(-gene_id) %>%
  dplyr::relocate(symbol)



ne.degs <- read.table(here("../Data/ChIPseq_NEboundgenes_208.txt")) %>% pull(x)
hdaci_mesc_degs <- read.csv(here("../Data/RNAseq_HDACi_DEGs_mESC.csv"), row.names = 1)
hdaci_mesc_tsatargets <- hdaci_mesc_degs %>% dplyr::filter(ASCL1_target %in% c("tsa", "both"))

counts.nepeaks <- counts %>% dplyr::filter(symbol %in% ne.degs)

mat <- counts.nepeaks %>%
  dplyr::select(symbol, contains("mESC") | contains("NE")) %>%
  tibble::column_to_rownames("symbol")
# Loop through genes of interest
genes.ofi <- ne.degs # c("Nkx6-1", "Dlx2", "Wnt5a", "Onecut1", "Plxna2", "Map2", "Hes5", "Tha1", "Phox2a", "Hdac11", "Nos1", "Sox10", "Ank2", "Dcx", "Ebf3", "Zbtb18")

hdaci_degs <- read.csv(here("../Data/RNAseq_HDACi_DEGs.csv"))

# define which padj columns map to which x-axis comparison
padj_map <- list(
  "mESC_None_DMSO|mESC_Dex_DMSO" = "mESC_ASCL1vsControl_padj",
  "mESC_None_TSA|mESC_Dex_TSA"   = "mESC_ASCL1withTSA_padj",
  "NE_None_DMSO|NE_Dex_DMSO"     = "NE_ASCL1vsControl_padj",
  "NE_None_TSA|NE_Dex_TSA"       = "NE_ASCL1withTSA_padj"
)

# x positions for each bracket (midpoint between the two conditions)
bracket_xpos <- c(1.5, 3.5, 5.5, 7.5)

for (gene.ofi in genes.ofi) {

  if (!gene.ofi %in% rownames(mat)) {
    message("Skipping ", gene.ofi, " - not found in counts matrix")
    next
  }

  counts.ofi <- mat[gene.ofi, ]

  df <- tibble::tibble(
    sample = names(counts.ofi),
    value  = as.numeric(counts.ofi)
  ) %>%
    separate(sample, into = c("exp", "line", "cell", "ASCL1", "TSA", "rep"), sep = "_") %>%
    mutate(
      color_group = case_when(
        cell == "mESC" & ASCL1 == "None" ~ "#AAE4E9",
        cell == "mESC" & ASCL1 == "Dex"  ~ "#4FAAB2",
        cell == "NE"   & ASCL1 == "None" ~ "#FFBCBD",
        cell == "NE"   & ASCL1 == "Dex"  ~ "#FB7D80"
      ),
      shape_group = ifelse(TSA == "DMSO", 21, 24),
      cond = paste(cell, ASCL1, TSA, sep = "_"),
      cond = factor(cond, levels = c(
        "mESC_None_DMSO", "mESC_Dex_DMSO", "mESC_None_TSA", "mESC_Dex_TSA",
        "NE_None_DMSO",   "NE_Dex_DMSO",   "NE_None_TSA",   "NE_Dex_TSA"
      ))
    )

  # get padj values for this gene from results table
  gene_row <- hdaci_degs %>% dplyr::filter(symbol == gene.ofi)

  padj_labels <- imap(padj_map, ~ {
      padj_val <- gene_row[[.x]]
      if (length(padj_val) == 0 || is.na(padj_val)) {
          return(tibble(comparison = .y, padj = NA, label = "ns"))
      }
      label <- case_when(
          padj_val < 0.001 ~ "***",
          padj_val < 0.01  ~ "**",
          padj_val < 0.05  ~ "*",
          TRUE             ~ "ns"
      )
      tibble(comparison = .y, padj = padj_val, label = label)
  }) %>% bind_rows()

  # set bracket height just above max value
  y_max    <- max(df$value, na.rm = TRUE)
  y_step   <- y_max * 0.08
  y_bracket <- y_max + y_step

  p <- ggplot(df, aes(x = cond, y = value)) +
    geom_boxplot(aes(fill = color_group), width = 0.7, outlier.shape = NA, color = "black", size = 0.3) +
    geom_point(aes(fill = color_group, shape = factor(shape_group)),
               position = position_jitter(width = 0.1, height = 0),
               size = 3, color = "black") +
    scale_fill_identity() +
    scale_shape_manual(values = c("21" = 21, "24" = 24)) +
    geom_vline(xintercept = 4.5, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    # add brackets and labels
    geom_segment(data = tibble(x = bracket_xpos - 0.4, xend = bracket_xpos + 0.4,
                               y = y_bracket, yend = y_bracket),
                 aes(x = x, xend = xend, y = y, yend = yend),
                 inherit.aes = FALSE, linewidth = 0.4) +
    annotate("text",
             x = bracket_xpos,
             y = y_bracket + y_step * 1.2,
             label = padj_labels$label,
             size = 7) +
    ylab("") +
    xlab("") +
    theme_cowplot() +
    theme(
      legend.position = "none",
      axis.line = element_line(linewidth = 0.4),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size = 14),
      axis.title.y = element_blank(),
      plot.title = element_text(size = 16, face = "italic", hjust = 0.5)
    ) +
    labs(title = gene.ofi)

  ggsave(
    filename = here("../Examples_HDACi/", paste0("Fig5D_", gene.ofi, ".png")),
    plot = p,
    width = 3, height = 4, units = "in", dpi = 300, bg = "white"
  )
}

```

Fig 5E - heatmap for genes that are upregulated by ascl1 relative to tsa baseline 

```{r}

library(ComplexHeatmap)
library(circlize)
library(here)
library(tidyverse)

counts <- read.csv(here("../Data/RNAseq_HDACi_normcounts.csv"))
counts <- counts %>% dplyr::rename(gene_id = X)

gene_map <- read.csv(here("../Data/RNAseq_GeneMap.csv"), row.names = 1)
counts <- counts %>%
  left_join(gene_map %>% dplyr::select(gene_id, symbol), by = "gene_id") %>%
  dplyr::filter(!is.na(symbol) & symbol != "") %>%
  dplyr::select(-gene_id) %>%
  dplyr::relocate(symbol) %>%
  group_by(symbol) %>%
  slice_max(order_by = rowMeans(across(where(is.numeric))), n = 1) %>%
  ungroup()

# ── Filter genes ──────────────────────────────────────────────────────────────
hdaci_degs <- read.csv(here("../Data/RNAseq_HDACi_DEGs.csv"))

genes_of_interest <- hdaci_degs %>%
  dplyr::filter(
    (mESC_ASCL1withTSA_padj < 0.01 & mESC_ASCL1withTSA_LFC > 0)
  ) %>%
  pull(symbol) %>%
  unique()


# ── Prepare matrix ────────────────────────────────────────────────────────────
mat_heatmap <- counts %>%
  dplyr::filter(symbol %in% genes_of_interest) %>%
  dplyr::select(symbol, contains("mESC")) %>%
  tibble::column_to_rownames("symbol")

# handle duplicate symbols if present
if (any(duplicated(rownames(mat_heatmap)))) {
  mat_heatmap <- mat_heatmap %>%
    rownames_to_column("symbol") %>%
    group_by(symbol) %>%
    slice_max(order_by = rowMeans(across(where(is.numeric))), n = 1) %>%
    ungroup() %>%
    column_to_rownames("symbol")
}

# z-score by row
mat_z <- t(scale(t(mat_heatmap)))

# ── Column annotation ─────────────────────────────────────────────────────────
col_meta <- tibble(sample = colnames(mat_z)) %>%
  separate(sample, into = c("exp", "line", "cell", "ASCL1", "TSA", "rep"), sep = "_") %>%
  mutate(
    condition = paste(ASCL1, TSA, sep = "_"),
    condition = factor(condition, levels = c("None_DMSO", "Dex_DMSO", "None_TSA", "Dex_TSA"))
  )

col_ann <- HeatmapAnnotation(
  ASCL1 = col_meta$ASCL1,
  TSA   = col_meta$TSA,
  col   = list(
    ASCL1 = c("None" = "#AAE4E9", "Dex"  = "#4FAAB2"),
    TSA   = c("DMSO" = "#D3D3D3", "TSA"  = "#404040")
  ),
  annotation_name_side = "left"
)

# ── Row annotation for NE-bound genes ────────────────────────────────────────
ne.degs <- read.table(here("../Data/ChIPseq_NEboundgenes_208.txt")) %>% pull(x)

ne_bound <- rownames(mat_z) %in% ne.degs

row_ann <- rowAnnotation(
  NE_bound = anno_mark(
    at        = which(ne_bound),
    labels    = rownames(mat_z)[ne_bound],
    labels_gp = gpar(fontsize = 16, fontface = "italic")
  )
)

# ── Draw heatmap ──────────────────────────────────────────────────────────────
ht <- Heatmap(
  mat_z,
  name              = "z-score",
  col = met.brewer("OKeeffe1", direction = -1),
  top_annotation    = col_ann,
  # right_annotation  = row_ann,
  column_order      = col_order,
  cluster_columns   = FALSE,
  cluster_rows      = TRUE,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  row_title         = paste0(length(genes_of_interest), " genes"),
  column_title      = "mESC",
  border            = TRUE
)

draw(ht)

png(here("../Figures/Fig_HDACi_Heatmap_mESC_TSAtargets.png"),
    width = 5, height = 7, units = "in", res = 300, bg = "white")
draw(ht)
dev.off()

```

Fig 5F - different heatmap to show genes that are originally NE specific but are now upregulated in mESC ascl1+tsa versus control

```{r}

library(ComplexHeatmap)
library(circlize)
library(here)
library(tidyverse)
library(MetBrewer)

counts <- read.csv(here("../Data/RNAseq_HDACi_normcounts.csv"))
counts <- counts %>% dplyr::rename(gene_id = X)

gene_map <- read.csv(here("../Data/RNAseq_GeneMap.csv"), row.names = 1)
counts <- counts %>%
  left_join(gene_map %>% dplyr::select(gene_id, symbol), by = "gene_id") %>%
  dplyr::filter(!is.na(symbol) & symbol != "") %>%
  dplyr::select(-gene_id) %>%
  dplyr::relocate(symbol) %>%
  group_by(symbol) %>%
  slice_max(order_by = rowMeans(across(where(is.numeric))), n = 1) %>%
  ungroup()

# ── Load gene lists ───────────────────────────────────────────────────────────
hdaci_degs <- read.csv(here("../Data/RNAseq_HDACi_DEGs.csv"))
ne.degs <- read.table(here("../Data/ChIPseq_NEboundgenes_208.txt")) %>% pull(x)

# genes significantly upregulated in ASCL1+TSA vs Control
ascl1tsa_up <- hdaci_degs %>%
  dplyr::filter(mESC_ASCL1withTSAvsControl_padj < 0.01 & mESC_ASCL1withTSAvsControl_LFC > 0) %>%
  pull(symbol)

# ── Prepare matrix for all ne.degs genes ─────────────────────────────────────
mat_heatmap <- counts %>%
  dplyr::filter(symbol %in% ne.degs) %>%
  dplyr::select(symbol, contains("mESC")) %>%
  tibble::column_to_rownames("symbol")

# z-score by row
mat_z <- t(scale(t(mat_heatmap)))

# ── Column annotation ─────────────────────────────────────────────────────────
col_meta <- tibble(sample = colnames(mat_z)) %>%
  separate(sample, into = c("exp", "line", "cell", "ASCL1", "TSA", "rep"), sep = "_") %>%
  mutate(
    condition = paste(ASCL1, TSA, sep = "_"),
    condition = factor(condition, levels = c("None_DMSO", "Dex_DMSO", "None_TSA", "Dex_TSA"))
  )

col_ann <- HeatmapAnnotation(
  ASCL1 = col_meta$ASCL1,
  TSA   = col_meta$TSA,
  col   = list(
    ASCL1 = c("None" = "#AAE4E9", "Dex" = "#4FAAB2"),
    TSA   = c("DMSO" = "#D3D3D3", "TSA" = "#404040")
  ),
  annotation_name_side = "left"
)

# ── Column order ──────────────────────────────────────────────────────────────
col_order <- order(col_meta$condition)

# ── Row annotation: label genes upregulated in ASCL1+TSA vs TSA alone ────────
ascl1tsa_vs_tsa_up <- hdaci_degs %>%
  dplyr::filter(mESC_ASCL1withTSA_padj < 0.01 & mESC_ASCL1withTSA_LFC > 0) %>%
  pull(symbol)

sig_up <- rownames(mat_z) %in% ascl1tsa_vs_tsa_up

row_ann <- rowAnnotation(
  ASCL1_TSA_up = anno_mark(
    at        = which(sig_up),
    labels    = rownames(mat_z)[sig_up],
    labels_gp = gpar(fontsize = 18, fontface = "italic")
  )
)

# ── Draw heatmap ──────────────────────────────────────────────────────────────
ht <- Heatmap(
  mat_z,
  name              = "z-score",
  col               = met.brewer("OKeeffe1", direction = -1),
  top_annotation    = col_ann,
  right_annotation  = row_ann,
  column_order      = col_order,
  cluster_columns   = FALSE,
  cluster_rows      = TRUE,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  row_title         = paste0(nrow(mat_z), " NE-bound genes"),
  column_title      = "mESC",
  border            = TRUE,
  km = 2
)

png(here("../Figures/Fig_HDACi_Heatmap_mESC_NEbound.png"),
    width = 7, height = 8, units = "in", res = 300, bg = "white")
draw(ht)
dev.off()

# ── Summary ───────────────────────────────────────────────────────────────────
n_total        <- nrow(mat_z)
n_ascl1tsa_vs_control <- sum(rownames(mat_z) %in% (hdaci_degs %>%
  dplyr::filter(mESC_ASCL1withTSAvsControl_padj < 0.01 & mESC_ASCL1withTSAvsControl_LFC > 0) %>%
  pull(symbol)))
n_tsa_alone    <- sum(rownames(mat_z) %in% (hdaci_degs %>%
  dplyr::filter(mESC_TSA_alone_padj < 0.01 & mESC_TSA_alone_LFC > 0) %>%
  pull(symbol)))
n_ascl1tsa_vs_tsa <- sum(sig_up)

message(
  "Of ", n_total, " NE-bound genes in the heatmap:\n",
  "  - ", n_ascl1tsa_vs_control, " upregulated in ASCL1+TSA vs Control\n",
  "  - ", n_tsa_alone,           " upregulated in TSA alone vs Control\n",
  "  - ", n_ascl1tsa_vs_tsa,     " upregulated in ASCL1+TSA vs TSA alone (labelled)"
)

```

Fig 5F' - GO analysis for DEGs with and without TSA

```{r, messages=FALSE}
library(here)
library(tidyverse)
library(clusterProfiler)
library(org.Mm.eg.db)
library(MetBrewer)
library(cowplot)

# ── GO analysis on TSA-sensitive ASCL1 targets (ns in DMSO, up in TSA) ───────
go_result <- enrichGO(
  gene          = genes_of_interest,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

go_simplified <- simplify(go_result)

p_go <- barplot(go_simplified, showCategory = 6) +
  theme_cowplot() +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
  scale_fill_gradient(high = "#e86f35", low = "#ffb945")

png(filename = here("../Figures/Fig_HDACi_GO_mESC_TSAsensitive.png"),
    width = 12, height = 12, units = "cm", res = 300, bg = "white")
p_go
dev.off()
```


# FIG S5 

Fig S5C-F - volcano plots for all comaprisons in supplementary and synergy comparison in mesc 

```{r}
library(ggplot2)
library(patchwork)
library(dplyr)
library(here)

hdaci_degs <- read.csv(file = here("../Data/RNAseq_HDACi_DEGs.csv"))

make_volcano <- function(data, lfc_col, padj_col, title,
                         lfc_thresh = 0, padj_thresh = 0.01,
                         up_colour = "#79c1d1") {
  df <- data %>%
    transmute(
      symbol        = symbol,
      LFC           = .data[[lfc_col]],
      padj          = .data[[padj_col]],
      neglog10_padj = -log10(padj),
      sig = case_when(
        padj < padj_thresh & LFC >  lfc_thresh ~ "up",
        padj < padj_thresh & LFC < -lfc_thresh ~ "down",
        TRUE                                   ~ "ns"
      ),
      sig = factor(sig, levels = c("ns", "down", "up"))
    )

  n_up   <- sum(df$sig == "up")
  n_down <- sum(df$sig == "down")

  ggplot(df, aes(x = LFC, y = neglog10_padj, colour = sig)) +
    geom_point(size = 1, alpha = 0.5) +
    annotate("text",
             x = Inf, y = Inf,
             hjust = 1.1, vjust = 1.5,
             label = paste0("Up: ", n_up),
             colour = up_colour, size = 3.5, fontface = "bold") +
    annotate("text",
             x = -Inf, y = Inf,
             hjust = -0.1, vjust = 1.5,
             label = paste0("Down: ", n_down),
             colour = "#404040", size = 3.5, fontface = "bold") +
    scale_colour_manual(
      values = c(up = up_colour, down = "#404040", ns = "#D3D3D3"),
      labels = c(up = "Up", down = "Down", ns = "NS")
    ) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    geom_hline(yintercept = -log10(padj_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    labs(
      title  = title,
      x      = "ashr Log2 Fold Change",
      y      = expression(-log[10](p[adj])),
      colour = NULL
    ) +
    theme_classic(base_size = 11) +
    scale_y_continuous(limits = c(0, 30)) +
    scale_x_continuous(limits = c(-8, 8)) +
    theme(
      plot.title      = element_text(hjust = 0.5, face = "bold"),
      legend.position = "bottom"
    )
}

# ── mESC ──────────────────────────────────────────────────────────────────────
mESC_p1 <- make_volcano(hdaci_degs, "mESC_ASCL1vsControl_ashrLFC",        "mESC_ASCL1vsControl_padj",        "ASCL1 vs Control",       up_colour = "#4FAAB2")
mESC_p2 <- make_volcano(hdaci_degs, "mESC_TSA_alone_ashrLFC",             "mESC_TSA_alone_padj",             "TSA only",               up_colour = "#4FAAB2")
mESC_p3 <- make_volcano(hdaci_degs, "mESC_ASCL1withTSAvsControl_ashrLFC", "mESC_ASCL1withTSAvsControl_padj", "ASCL1 + TSA vs Control", up_colour = "#4FAAB2")
mESC_p4 <- make_volcano(hdaci_degs, "mESC_ASCL1withTSA_ashrLFC",          "mESC_ASCL1withTSA_padj",          "ASCL1 with TSA",         up_colour = "#4FAAB2")

mESC_full <- mESC_p1 + mESC_p2 + mESC_p3 + mESC_p4 +
  plot_layout(ncol = 4, guides = "collect") +
  plot_annotation(title = "mESC") &
  theme(legend.position = "bottom")

ggsave(mESC_full, filename = here("../Figures/Fig_HDACi_RNAseq_mESC_new.png"), width = 12, height = 4, dpi = 300)

# ── NE ────────────────────────────────────────────────────────────────────────
NE_p1 <- make_volcano(hdaci_degs, "NE_ASCL1vsControl_ashrLFC",        "NE_ASCL1vsControl_padj",        "ASCL1 vs Control",       up_colour = "#FB7D80")
NE_p2 <- make_volcano(hdaci_degs, "NE_TSA_alone_ashrLFC",             "NE_TSA_alone_padj",             "TSA only",               up_colour = "#FB7D80")
NE_p3 <- make_volcano(hdaci_degs, "NE_ASCL1withTSAvsControl_ashrLFC", "NE_ASCL1withTSAvsControl_padj", "ASCL1 + TSA vs Control", up_colour = "#FB7D80")
NE_p4 <- make_volcano(hdaci_degs, "NE_ASCL1withTSA_ashrLFC",          "NE_ASCL1withTSA_padj",          "ASCL1 with TSA",         up_colour = "#FB7D80")

NE_full <- NE_p1 + NE_p2 + NE_p3 + NE_p4 +
  plot_layout(ncol = 4, guides = "collect") +
  plot_annotation(title = "NE") &
  theme(legend.position = "bottom")

ggsave(NE_full, filename = here("../Figures/Fig_HDACi_RNAseq_NE_new.png"), width = 12, height = 4, dpi = 300)

# ── mESC synergy ──────────────────────────────────────────────────────────────
mESC_synergy <- make_volcano(hdaci_degs, "mESC_synergy_ashrLFC", "mESC_synergy_padj", "mESC Synergy", up_colour = "#4FAAB2")

ggsave(mESC_synergy, filename = here("../Figures/Fig_HDACi_RNAseq_mESC_synergy_new.png"), width = 4, height = 4, dpi = 300)

# ── NE synergy ────────────────────────────────────────────────────────────────
NE_synergy <- make_volcano(hdaci_degs, "NE_synergy_ashrLFC", "NE_synergy_padj", "NE Synergy", up_colour = "#FB7D80")

ggsave(NE_synergy, filename = here("../Figures/Fig_HDACi_RNAseq_NE_synergy_new.png"), width = 4, height = 4, dpi = 300)

```


# FIG 6

Fig 6A - cofactor motif versus expression for NE 

```{r, messages=FALSE}
library(here)
library(tidyverse)
library(AnnotationHub)
library(DESeq2)
library(ggrepel)

# Load and clean HOMER motif results
homer.ne <- read.delim(
  here("../Data/homer/ne_specific/knownResults.txt"),
  header = TRUE,
  comment.char = "",
  quote = ""
) %>%
  extract(
    Motif.Name,
    into = c("TF", "family"),
    regex = "^([^\\(]+)\\(([^\\)]+)\\)",
    remove = FALSE
  ) %>%
  mutate(
    target_pct      = as.numeric(str_remove(X..of.Target.Sequences.with.Motif, "%")),
    background_pct  = as.numeric(str_remove(X..of.Background.Sequences.with.Motif, "%")),
    fold_enrichment = target_pct / background_pct,
    logPvalue       = -log2(P.value),
    TF_family_clean = TF %>%
      str_replace_all("\\.", "-") %>%
      str_to_sentence()
  ) %>%
  dplyr::select(
    TF_family_clean,
    TF,
    family,
    Consensus,
    P.value,
    target_pct,
    background_pct,
    fold_enrichment,
    logPvalue
  )

# Load reference TF list

tfs <- read.csv(here("../Data/Ref_MouseTFs.csv"))

homeoboxes <- tfs %>%
  dplyr::filter(Family == "Homeobox") %>%
  dplyr::pull(Gene.Symbol)

tfs.clean <- tfs %>%
  dplyr::pull(Gene.Symbol)

# Load DESeq2 object and map Ensembl IDs to symbols
dds <- readRDS(here("../Data/RNAseq_ASCL1_dds.Rds"))

ah <- AnnotationHub()
EnsDb <- ah[["AH109367"]]

gene_map <- genes(EnsDb, return.type = "DataFrame") %>%
  as.data.frame() %>%
  dplyr::select(gene_id, symbol)

gene_symbols <- gene_map$symbol[match(rownames(dds), gene_map$gene_id)]
dds <- dds[!is.na(gene_symbols) & gene_symbols != "", ]
rownames(dds) <- gene_symbols[!is.na(gene_symbols) & gene_symbols != ""]

# Differential expression: mESC vs NE
res <- results(
  dds,
  contrast = c("Group", "gG5_NE_t0", "gG5_mESC_t0"),
  alpha = 0.01,
  lfcThreshold = 0.585
)

lfc <- lfcShrink(dds, res = res, type = "ashr")

sig.tfs <- cbind(
  as.data.frame(res),
  ashrLFC = lfc$log2FoldChange,
  ashrSE  = lfc$lfcSE,
  symbol  = rownames(res)
) %>%
  dplyr::filter(
    symbol %in% tfs.clean,
    log2FoldChange > 0,
    padj < 0.01
  )

# Merge motif enrichment with TF expression
final.tbl <- homer.ne %>%
  inner_join(sig.tfs, by = c("TF_family_clean" = "symbol")) %>%
  mutate(
    motif_class = case_when(
      family == "HMG"       ~ "HMG",
      family == "Homeobox" ~ "Homeobox",
      family == "bHLH" ~ "bHLH",
      TRUE                 ~ "Other"
    )
  )

final.tbl <- final.tbl %>%
  mutate(
    label_tf = ifelse(
      padj < 0.05 & P.value < 0.05,
      TF_family_clean,
      NA_character_
    )
  )

# Plot: Motif enrichment vs TF expression
plot <- ggplot(final.tbl, aes(x = fold_enrichment, y = ashrLFC)) +
  geom_point(
    aes(color = motif_class),
    size = 3,
    alpha = 0.8
  ) +
  scale_color_manual(
    values = c(
      HMG = "orange",
      Homeobox = "magenta",
      bHLH = "lightblue",
      Other = "grey70"
    )
  ) +
  geom_text_repel(
    aes(label = label_tf),
    size = 5,
    max.overlaps = 20,
    box.padding = 0.5
  ) +
  labs(
    x = "Motif fold enrichment (NE vs background)",
    y = "RNA-seq log2 fold change (NE vs mESC)",
    color = "Motif family"
  ) +
  theme_classic()

plot

ggsave(filename = here("../Figures/Fig6A.png"), width = 5, height = 5, bg = "white", dpi = 300)

```


# FIG S6

Fig S6A - cofactor motif versus expression for mESC 

```{r, messages=FALSE}
library(here)
library(tidyverse)
library(AnnotationHub)
library(DESeq2)
library(ggrepel)

# -----------------------------
# Load and clean HOMER motif results (mESC-specific)
# -----------------------------
homer.mesc <- read.delim(
  here("../Data/homer/mesc_specific/knownResults.txt"),
  header = TRUE,
  comment.char = "",
  quote = ""
) %>%
  extract(
    Motif.Name,
    into = c("TF", "family"),
    regex = "^([^\\(]+)\\(([^\\)]+)\\)",
    remove = FALSE
  ) %>%
  mutate(
    target_pct      = as.numeric(str_remove(X..of.Target.Sequences.with.Motif, "%")),
    background_pct  = as.numeric(str_remove(X..of.Background.Sequences.with.Motif, "%")),
    fold_enrichment = target_pct / background_pct,
    logPvalue       = -log2(P.value),
    TF_family_clean = TF %>%
      str_replace_all("\\.", "-") %>%
      str_to_sentence(),
    TF_family_clean = dplyr::case_when(
      TF_family_clean == "Ap-2alpha" ~ "Tfap2a",
      TF_family_clean == "Ap-2gamma" ~ "Tfap2c",
      TRUE                           ~ TF_family_clean
    )
  ) %>%
  dplyr::select(
    TF_family_clean,
    TF,
    family,
    Consensus,
    P.value,
    target_pct,
    background_pct,
    fold_enrichment,
    logPvalue
  )

# -----------------------------
# Load reference TF list
# -----------------------------
tfs <- read.csv(here("../Data/Ref_MouseTFs.csv"))

tfs.clean <- tfs %>%
  dplyr::pull(Gene.Symbol)

# -----------------------------
# Load DESeq2 object and map Ensembl IDs to symbols
# -----------------------------
dds <- readRDS(here("../Data/RNAseq_ASCL1_dds.Rds"))

ah <- AnnotationHub()
EnsDb <- ah[["AH109367"]]

gene_map <- genes(EnsDb, return.type = "DataFrame") %>%
  as.data.frame() %>%
  dplyr::select(gene_id, symbol)

gene_symbols <- gene_map$symbol[match(rownames(dds), gene_map$gene_id)]
dds <- dds[!is.na(gene_symbols) & gene_symbols != "", ]
rownames(dds) <- gene_symbols[!is.na(gene_symbols) & gene_symbols != ""]

# -----------------------------
# Differential expression: mESC vs NE
# (negative LFC = higher in mESC)
# -----------------------------
res <- results(
  dds,
  contrast = c("Group", "gG5_NE_t0", "gG5_mESC_t0"),
  alpha = 0.01,
  lfcThreshold = 0.585
)

lfc <- lfcShrink(dds, res = res, type = "ashr")

sig.tfs.mesc <- cbind(
  as.data.frame(res),
  ashrLFC = lfc$log2FoldChange,
  ashrSE  = lfc$lfcSE,
  symbol  = rownames(res)
) %>%
  dplyr::filter(
    symbol %in% tfs.clean,
    ashrLFC < 0,
    padj < 0.01
  )

# -----------------------------
# Merge motif enrichment with TF expression
# -----------------------------
final.tbl.mesc <- homer.mesc %>%
  inner_join(sig.tfs.mesc, by = c("TF_family_clean" = "symbol")) %>%
  mutate(
    motif_class = case_when(
      family == "AP2"       ~ "AP2",
      family == "NR" ~ "NR",
      family == "Zf"     ~ "Zf",
      family == "Homeobox"     ~ "Homeobox",
      TRUE                 ~ "Other"
    ),
    label_tf = ifelse(
      padj < 0.05 & P.value < 0.05,
      TF_family_clean,
      NA_character_
    )
  )

# -----------------------------
# Plot: Motif enrichment vs TF expression (mESC)
# -----------------------------
plot <- ggplot(final.tbl.mesc, aes(x = fold_enrichment, y = -ashrLFC)) +
  geom_point(
    aes(color = motif_class),
    size = 3,
    alpha = 0.8
  ) +
  scale_color_manual(
    values = c(
      AP2 = "green",
      Homeobox = "magenta",
      NR = "yellow",
      Zf = "cyan",
      Other = "grey70"
    )
  ) +
  geom_text_repel(
    aes(label = label_tf),
    size = 5,
    max.overlaps = 20,
    box.padding = 0.5,
    na.rm = TRUE
  ) +
  labs(
    x = "Motif fold enrichment (mESC vs background)",
    y = "RNA-seq log2 fold change (NE vs mESC)",
    color = "Motif family"
  ) +
  theme_classic()

ggsave(
  filename = here("../Figures/FigS6A.png"),
  plot = plot,
  width = 6,
  height = 5,
  bg = "white",
  dpi = 300
)

plot

```


# FIG 7

Fig 7A-A'' - heatmap for all degs in mesc with ascl1 and phox2b 

```{r}
library(here)
library(tidyverse)
library(ComplexHeatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(cowplot)
library(MetBrewer)

# which degs to look at? all ascl1 + phox2b genes 

phox2b_degs <- read.table(file = here("../Data/RNAseq_Cofactor_Phox2b_DEGs.csv"), sep = ",", header = TRUE)

phox2bascl1mescgenes <- phox2b_degs %>% dplyr::filter(mESC_Combo_change == "up" & mESC_Combo_LFC > 1) %>% pull(symbol) %>% unique() # 757 genes in this list, which matches the volcano 

# ── Data loading ──────────────────────────────────────────────────────────────
counts.phox2b <- read.table(
  file = here("../Data/RNAseq_Cofactors_Phox2b_normcounts.csv"),
  sep = ",", header = TRUE
) %>% dplyr::rename(symbol = X)


# ── Filter to upregulated genes from combined condition ─────────────────────────────────────

counts.phox2bdegs <- counts.phox2b %>%
  dplyr::filter(symbol %in% phox2bascl1mescgenes)

counts.phox2bdegs.clean <- counts.phox2bdegs %>%
  dplyr::filter(symbol != "" & !is.na(symbol)) %>%
  distinct(symbol, .keep_all = TRUE)

# ── Build z-scored expression matrix ─────────────────────────────────────────

expr_mat <- counts.phox2bdegs.clean %>%
  dplyr::select(symbol, contains("mESC")) %>%
  column_to_rownames("symbol") %>%
  as.matrix()
expr_z <- t(scale(t(expr_mat)))


# ── Cluster rows BEFORE heatmap (set seed for reproducibility) ───────────────
set.seed(42)
km_result           <- kmeans(expr_z, centers = 2, nstart = 25)
cluster_assignments <- km_result$cluster

cluster1_genes <- names(cluster_assignments[cluster_assignments == 1])
cluster2_genes <- names(cluster_assignments[cluster_assignments == 2])
#cluster3_genes <- names(cluster_assignments[cluster_assignments == 3])
#cluster2and3_genes <- c(cluster2_genes, cluster3_genes)
# etc. 

# ── Row split factor ──────────────────────────────────────────────────────────
row_split <- factor(paste0("C", cluster_assignments), levels = paste0("C", 1:2))

# # ── Right annotation: gene labels ────────────────────────────────────────────
# label_genes <- head(genes_of_interest, 10)
# row_labels  <- rownames(expr_z_withNE)
# 
# ha_right <- rowAnnotation(
#   link = anno_mark(
#     at        = which(row_labels %in% label_genes),
#     labels    = row_labels[which(row_labels %in% label_genes)],
#     labels_gp = gpar(fontsize = 8),
#     padding   = unit(1, "mm")
#   )
# )


# ── Heatmap ───────────────────────────────────────────────────────────────────
hm <- Heatmap(
  expr_z,
  name               = "Z-score",
  show_row_names     = FALSE,
  show_column_names  = FALSE,
  cluster_rows       = TRUE,
  cluster_columns    = FALSE,
  row_split          = row_split,
  cluster_row_slices = FALSE,
  row_title          = "Phox2b-associated genes",
  column_title       = "Samples",
  col = met.brewer("OKeeffe1", direction = -1)
)

draw(hm)

png(
  filename = here("../Figures/Fig7A.png"),
  width = 9, height = 24, units = "cm", bg = "white", res = 300
)
draw(hm)
dev.off()



# ── GO analysis on target cluster genes ──────────────────────────────────────
go_result_1 <- enrichGO(
  gene          = cluster1_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

go_result_2 <- enrichGO(
  gene          = cluster2_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)


# ── Simplify and plot GO results for each cluster ─────────────────────────────
tam_cols <- MetBrewer::met.brewer("Tam", direction = -1)

plot_go <- function(go_result, cluster_label, fig_filename, csv_filename) {
  if (!is.null(go_result) && nrow(as.data.frame(go_result)) > 0) {
    
    go_simplified <- clusterProfiler::simplify(go_result, cutoff = 0.7)
    go_df         <- as.data.frame(go_simplified)
    print(head(go_df))
    
    write.csv(go_df, file = here(paste0("../Data/", csv_filename, ".csv")), row.names = FALSE)
    
    p <- dotplot(go_simplified, showCategory = 6) +
      scale_fill_gradientn(colours = tam_cols) +
      scale_y_discrete(labels = function(x) str_wrap(x, width = 45)) +
      ggtitle(paste0("GO:BP — Cluster ", cluster_label)) +
      theme_cowplot()
    
    print(p)
    
    ggsave(
      filename = here(paste0("../Figures/", fig_filename, ".png")),
      plot     = p,
      width    = 15,
      height   = 9,
      units    = "cm",
      dpi      = 300,
      bg       = "white"
    )
    
  } else {
    cat("No significant GO terms found for Cluster", cluster_label, "\n")
  }
}

plot_go(go_result_1, "1", "Fig7A_GO_C1", "cluster1_GO")
plot_go(go_result_2, "2", "Fig7A_GO_C2", "cluster2_GO")
# plot_go(go_result_3, "3", "Fig7A_GO_C3", "cluster3_GO")
# plot_go(go_result_2and3, "2and3", "Fig7A_GO_C23", "cluster2and3_GO")

# ── Save gene lists ───────────────────────────────────────────────────────────
write.csv(data.frame(symbol = cluster1_genes), file = here("../Data/cluster1_genes.csv"), row.names = FALSE)
write.csv(data.frame(symbol = cluster2_genes), file = here("../Data/cluster2_genes.csv"), row.names = FALSE)
#write.csv(data.frame(symbol = cluster3_genes), file = here("../Data/cluster3_genes.csv"), row.names = FALSE)
```

Fig 7B-B' - previously identified mESC specific genes incl. cluster that go down with phox2b 

```{r}
library(here)
library(tidyverse)
library(ComplexHeatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(MetBrewer)
library(cowplot)

# ── Data loading ──────────────────────────────────────────────────────────────
counts.phox2b <- read.table(
  file = here("../Data/RNAseq_Cofactors_Phox2b_normcounts.csv"),
  sep = ",", header = TRUE
) %>% dplyr::rename(symbol = X)

degs <- read.csv(file = here("../Data/RNAseq_ASCL1_DEGs.csv"))

# ── Filter to mESC-only upregulated genes ────────────────────────────────────
mesconly <- degs %>%
  dplyr::filter(
    NE_t24_gg5_change    == "ns",
    EpiLC_t24_gg5_change == "ns",
    mESC_t24_gg5_change  == "up"
  ) %>%
  pull(symbol) %>%
  unique()

counts.phox2bmescgenes <- counts.phox2b %>%
  dplyr::filter(symbol %in% mesconly)

counts.phox2bdegs.clean <- counts.phox2bmescgenes %>%
  dplyr::filter(symbol != "" & !is.na(symbol)) %>%
  distinct(symbol, .keep_all = TRUE)

# ── Build z-scored expression matrix ─────────────────────────────────────────
expr_mat <- counts.phox2bdegs.clean %>%
  dplyr::select(symbol, contains("mESC")) %>%
  column_to_rownames("symbol") %>%
  as.matrix()

expr_z <- t(scale(t(expr_mat)))

# ── Cluster rows BEFORE heatmap (set seed for reproducibility) ───────────────
set.seed(42)
km_result <- kmeans(expr_z, centers = 3, nstart = 25)
cluster_assignments <- km_result$cluster


# ── Identify column groups ────────────────────────────────────────────────────
col_names <- colnames(expr_z)

ascl1_alone_cols  <- col_names[grepl("Ascl1_NoDox", col_names)]
phox2b_ascl1_cols <- col_names[grepl("Ascl1_Dox",   col_names)]

cat("Ascl1 alone columns:\n");  print(ascl1_alone_cols)
cat("Phox2b+Ascl1 columns:\n"); print(phox2b_ascl1_cols)

# ── Calculate per-cluster means and find target cluster ───────────────────────
cluster_means <- data.frame(
  cluster     = 1:3,
  ascl1_mean  = tapply(rowMeans(expr_z[, ascl1_alone_cols,  drop = FALSE]), cluster_assignments, mean),
  phox2b_mean = tapply(rowMeans(expr_z[, phox2b_ascl1_cols, drop = FALSE]), cluster_assignments, mean)
) %>%
  mutate(phox2b_lower = phox2b_mean < ascl1_mean)

print(cluster_means)

target_cluster <- cluster_means %>%
  dplyr::filter(phox2b_lower) %>%
  mutate(diff = ascl1_mean - phox2b_mean) %>%
  slice_max(diff, n = 1) %>%
  pull(cluster)

cat("Target cluster:", target_cluster, "\n")

genes_of_interest <- names(cluster_assignments[cluster_assignments == target_cluster])
cat("Number of genes:", length(genes_of_interest), "\n")

# ── Build heatmap using pre-computed clusters (reproducible) ──────────────────
row_split <- factor(paste0("C", cluster_assignments), levels = paste0("C", 1:3))

# Label only the first 10 genes of interest
label_genes <- head(genes_of_interest, 10)
row_labels   <- rownames(expr_z)
ha_right <- rowAnnotation(
  link = anno_mark(
    at    = which(row_labels %in% label_genes),
    labels = row_labels[which(row_labels %in% label_genes)],
    labels_gp = gpar(fontsize = 8),
    padding = unit(1, "mm")
  )
)

hm <- Heatmap(
  expr_z,
  name              = "Z-score",
  show_row_names    = FALSE,
  show_column_names = FALSE,
  cluster_rows      = TRUE,
  cluster_columns   = FALSE,
  row_split         = row_split,
  cluster_row_slices = FALSE,
  row_title         = "Phox2b-associated genes",
  column_title      = "Samples",
  col = met.brewer("OKeeffe1", direction = -1)
)

draw(hm)

png(
  filename = here("../Figures/Fig6H-1.png"),
  width = 9, height = 24, units = "cm", bg = "white", res = 300
)
draw(hm)
dev.off()

# ── Check gene symbol mapping before GO ──────────────────────────────────────
test_map <- clusterProfiler::bitr(
  head(genes_of_interest, 10),
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Mm.eg.db
)
print(test_map)

# ── GO analysis on target cluster genes ──────────────────────────────────────
go_result <- enrichGO(
  gene          = genes_of_interest,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

if (!is.null(go_result) && nrow(as.data.frame(go_result)) > 0) {
  go_simplified <- clusterProfiler::simplify(go_result, cutoff = 0.7)
  go_df <- as.data.frame(go_simplified)
  print(head(go_df))
  write.csv(go_df, file = here("../Data/cluster_phox2b_lower_GO.csv"), row.names = FALSE)
  tam_cols <- MetBrewer::met.brewer("Tam", direction = -1)
  p <- dotplot(go_simplified, showCategory = 6) +
    scale_fill_gradientn(colours = tam_cols) +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 20)) +
    ggtitle("GO:BP — mESC genes lower in Phox2b+Ascl1") +
    theme_cowplot()
  print(p)
  ggsave(
    filename = here("../Figures/Fig6I_GO-2.png"),
    plot     = p,
    width    = 12,
    height   = 14,
    units    = "cm",
    dpi      = 300,
    bg       = "white"
  )
} else {
  cat("No significant GO terms found\n")
}

# ── Save gene list ────────────────────────────────────────────────────────────
write.csv(
  data.frame(symbol = genes_of_interest),
  file = here("../Data/cluster_phox2b_lower_genes.csv"),
  row.names = FALSE
)

```

Fig 7F-G - quantifying TUBB3-positive cells 

```{r, messages=FALSE}

library(dplyr)
library(ggplot2)
library(ggpubr)
library(here)

data <- read.csv(file = here("Nuclei_2.csv"))

data <- data %>% mutate(Metadata_Condition = recode(Metadata_Condition, "dex" = "ascl1"))

tubb3_percentage <- data %>%
  filter(!is.na(Metadata_Celltype) & Metadata_Celltype != "") %>%
  filter(Metadata_Cofactor != "gbx1" & Metadata_Cofactor != "gg5") %>%
  group_by(Metadata_Image, Metadata_Celltype, Metadata_Condition, Metadata_Cofactor, Metadata_Rep, Metadata_Well) %>%
  summarise(
    Total_Cells = n(),
    Positive_Cells = sum(Intensity_MeanIntensity_Tubb3 > 0.16),
    Percent_Positive = (Positive_Cells / Total_Cells) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    Metadata_Celltype = factor(Metadata_Celltype, levels = c("mESC", "EpiLC", "NE")),
    Metadata_Condition = factor(Metadata_Condition, levels = c("control", "ascl1", "dox", "ascl1dox")),
    Metadata_Cofactor = factor(Metadata_Cofactor, levels = c("gg5", "dlx3", "phox2a", "phox2b", "lmx1a", "gbx1"))
  )

comparisons <- list(
  c("ascl1", "control"),
  c("ascl1", "dox"),
  c("ascl1", "ascl1dox")
)

# Define colors per cell type
colors_list <- list(
  "mESC"  = c("control" = "#AAE4E9", "dox" = "#AAE4E9", "ascl1" = "#4FAAB2", "ascl1dox" = "#4FAAB2"),
  "EpiLC" = c("control" = "#FFE8B9", "dox" = "#FFE8B9", "ascl1" = "#FFD170", "ascl1dox" = "#FFD170"),
  "NE"    = c("control" = "#FFBCB9", "dox" = "#FFBCB9", "ascl1" = "#FF7770", "ascl1dox" = "#FF7770")
)

plots <- lapply(levels(tubb3_percentage$Metadata_Celltype), function(ct) {
  df <- filter(tubb3_percentage, Metadata_Celltype == ct)
  ggplot(df, aes(x = Metadata_Condition, y = Percent_Positive, fill = Metadata_Condition)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.7, color = "black") +
    facet_wrap(~Metadata_Cofactor, nrow = 1) +
    scale_fill_manual(values = colors_list[[ct]]) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = ct, x = "Condition", y = "Percentage of Tubb3-positive cells") +
    stat_compare_means(
      comparisons = comparisons,
      method = "t.test",
      label = "p.signif",
      p.adjust.method = "BH",
      tip.length = 0.01,
      step.increase = 0.1,
      aes(group = Metadata_Condition),label.y = 25
    )
})

# Save each plot
for (i in seq_along(plots)) {
  ggsave(plots[[i]], filename = here(paste0(levels(tubb3_percentage$Metadata_Celltype)[i], "_tubb3_plot.png")),
         width = 10, height = 3)
}

plots

```


# FIG S7 

Fig S7A-A'' - volcano plots for all conditions in mesc 

```{r}

phox2b_degs <- read.table(file = here("../Data/RNAseq_Cofactor_Phox2b_DEGs.csv"), sep = ",", header = TRUE)

library(ggplot2)
library(patchwork)
library(ggrepel)

make_volcano <- function(data, lfc_col, padj_col, title,
                         lfc_thresh = 1, padj_thresh = 0.01) {
  df <- data %>%
    transmute(
      symbol,
      LFC           = .data[[lfc_col]],
      padj          = .data[[padj_col]],
      neglog10_padj = -log10(padj),
      sig = case_when(
        padj < padj_thresh & LFC >  lfc_thresh ~ "up",
        padj < padj_thresh & LFC < -lfc_thresh ~ "down",
        TRUE                                   ~ "ns"
      ),
      sig = factor(sig, levels = c("ns", "down", "up"))
    )
  
  # ── Count up/down for annotation ───────────────────────────────────────────
  n_up   <- sum(df$sig == "up")
  n_down <- sum(df$sig == "down")
  
  ggplot(df, aes(x = LFC, y = neglog10_padj, colour = sig)) +
    geom_point(size = 1, alpha = 0.5) +
    annotate("text",
             x = Inf, y = Inf,
             hjust = 1.1, vjust = 1.5,
             label = paste0("Up: ", n_up),
             colour = "#79c1d1", size = 3.5, fontface = "bold") +
    annotate("text",
             x = -Inf, y = Inf,
             hjust = -0.1, vjust = 1.5,
             label = paste0("Down: ", n_down),
             colour = "#404040", size = 3.5, fontface = "bold") +
    scale_colour_manual(
      values = c(
        up   = "#79c1d1",
        down = "#404040",
        ns   = "#D3D3D3"
      ),
      labels = c(up = "Up", down = "Down", ns = "NS")
    ) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    geom_hline(yintercept = -log10(padj_thresh),
               linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    labs(
      title  = title,
      x      = "Log2 Fold Change",
      y      = expression(-log[10](p[adj])),
      colour = NULL
    ) +
    theme_classic(base_size = 11) +
    scale_y_continuous(limits = c(0, 300)) +
    scale_x_continuous(limits = c(-6, 8)) +
    theme(
      plot.title      = element_text(hjust = 0.5, face = "bold"),
      legend.position = "bottom"
    )
}

# ── Build each plot ───────────────────────────────────────────────────────────
p1 <- make_volcano(phox2b_degs, "mESC_Ascl1_LFC",  "mESC_Ascl1_padj",  "Control vs ASCL1")
p2 <- make_volcano(phox2b_degs, "mESC_Phox2b_LFC", "mESC_Phox2b_padj", "Control vs PHOX2B")
p3 <- make_volcano(phox2b_degs, "mESC_Combo_LFC",  "mESC_Combo_padj",  "Control vs ASCL1+PHOX2B")

# ── Combine ───────────────────────────────────────────────────────────────────
full <- p1 + p2 + p3 + plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")

ggsave(full, filename = here("../Figures/Fig_Phox2b_RNAseq.png"), width = 9, height = 4, dpi = 300)

```

